
# **Lab 2 data manipulation and presentation in R**{-}


## **Learning objectives**{-} 

<details>
  <summary>Click to expand</summary>
__By the end of this lab you will be able to:__


*	Demonstrate the ability to import and export variety of data types

*	Demonstrate the ability to transform and manipulate data using a variety of base and accessory functions

    + Know haw to manipulate data using functions from the tidyvesre family of packages e.g. dplyr and tidyr
  
*	Demonstrate the ability to use and nest conditional statements

*	Understand the concept of iteration and demonstrate the ability to iterate over complex data

*	Demonstrate the ability to write costume functions in R

*	Demonstrate the ability to generate plots in R

    + Demonstrate mastery of plotting using the ggplot package
  
</details>


## **Setup**{-}


As part of this lab you will learn how to import, export, manipulate and present data in R.
You will also learn how to use loops and conditional statements to control the flow of your script.
Lastly you will learn how to write costume (user-defined) functions. 

As part of this lab you will use some packages belonging to the [Tidyverse](https://www.tidyverse.org) package family (the 
link has more useful information). Tydiverse packages are geared towards data importation, manipulation, and presentation. 
They all use standardized grammar and philosophy that make the code more streamlined and readable. Stringer, which you have
already used, is part of the Tidyverse packages. 

It is possible to install packages belonging to the tidyverse individually, however all Tidyverse packages can also be 
installed using a wrapper package called...tidyverse. Before you start this lab, use the code bellow to install and load 
the tidyverse package. 
Installing the 

```{r tidyverse install, eval=F, echo=T}
install.packages("tidyverse")
library("tidyverse")

```

```{r load packages, echo=FALSE,message=FALSE,warning=FALSE,include=FALSE}
library("tidyverse")
library("DT")

```

***

>**Note: even if you installed all tidyverse packages using `install.packages("tidyverse")`, tidyverse packages such as** 
>**`dplyr` can still be loaded individually**

***

## **Importing and exporting data in R**{-}

So far you have been generating data in R, however when you work on real data you will need to import it to R. 
R is capable of importing multiple file formats, and additional packages can expand its capabilities to import specialized 
file formats (e.g. raw DNA and RNA sequencing files).
During this lab you will import csv and txt files, make sure to download the CSV and txt files from Brightspace.

### **read.csv**{-}

[CSV files](https://en.wikipedia.org/wiki/Comma-separated_values) (or comma-separated value files) are text files where 
values are separated by commas. This is a commonly used file format used to store tabulated data. 
`read.csv()` is a built-in function is base R that imports CSV files to R as a data frame. Note that the `readr` package 
which is part of the Tidyverse has a `read_csv()` function which imports CSV files to R.
However, it imports them as a tibble which is a sort of a redesigned data frame with some quirks so make sure not to
confuse them.

```{r read.csv 1}

#When calling read.csv, the file arguments specifies the path the the file you are loading 
#/path/to/file.csv
#eg in windows c:/directory1/directory2/file.csv
#eg in mac or linux ~/directory1/file.csv

df1 <- read.csv(file ="lab2_files/practice2.csv")
class(df1)
head(df1,3)
#Note the column with the x, that's the column with the row names.
#If you know that the first column in file you are loading conatins row names add
#row.names = 1 as part of the read.csv function

#Note that read.csv import characters as factors, to avoid it use stringsAsFactors=F
class(df1$X)


#Let's fix that by assigning 
row.names(df1) <- df1$X
#Now let's remove the first column
df1 <- df1[,-1]
#Another way to do this is 
#df1[1] <- NULL
#Could use $ as well
head(df1)

```

`read.csv()` can be used to read files from online sources as well, the code snippet bellow demonstrates how to read a file
from a public repository on my GitHub page. 

```{r read.csv2}

df2 <- read.csv(
  file = "https://raw.githubusercontent.com/idohatam/Biol-3315-files/main/practice.csv",
  row.names = 1, stringsAsFactors = FALSE)
head(df2,3)

```


### **read.table**{-}

`read.csv()` is a private case of the `read.table()` function (R has 
[several](https://www.rdocumentation.org/packages/utils/versions/3.6.2/topics/read.table) of those. `read.table()` allows 
you to read tabulated data stored in text files, but separated by any type of spacer.

Here is an example where the spacer is a tab. Note that read.table can be used to read csv files as well and that read.csv 
can be used to read txt files so long as the sep argument is correct.

```{r read.table1}

#sep = '\t' means tab delimited
df1b <- read.table("lab2_files/practice2.txt", sep = '\t', header = T, 
                   row.names = 1, stringsAsFactors = F)

class(df1b)

#are they identical?
df1 == df1b


```

***

>**A note about funky files and funky R behavior**
>
>As bioinformaticians, you will be asked to analyze data that you did not generate. If the columns contain 
>characters that R recognizes as special characters, it might lead to errors in the delimitation process when importing the
>file. `read.table()` and all private cases of it, has a quote argument that help circumvent these errors by ignoring 
specified characters. 
This [blog post](https://biowize.wordpress.com/2013/10/08/quotation-marks-and-rs-read-table-function/) has a great example 
>of how to use the quote argument in `read.table()`.

***

### **Reading non-tabulated text files**{-}

As bioinformaticians you will be required to work with non-tabulated file formats. A great example of this are text files, 
where information is stored as lines of strings. The [FASTA file format](https://en.wikipedia.org/wiki/FASTA_format), a 
standard for handling sequence data (will be covered in labs 3 and 4) is in fact a text file with lines as strings of 
sequences.
Base R has some built-in functions to import non-tabulated text files. The snippet bellow demonstrates how read a text file
from the Gutenberg organisation. Just like with `read.csv()`, you can use it to read files locally as well.

```{r non-tabulated txt file}

#The function file establishes a connection with the file, open="r" denotes a read only connection
book <- file("https://www.gutenberg.org/files/98/98-0.txt", open="r")

#The function readLines read the file to memory line by line
#This is also a useful way to remove non tabulated lines from a file that otherwise contains tabulated data (e.g. header lines)

book.lines <- readLines(book)

#The file is stored as a large atomic vector

typeof(book.lines)
is.atomic(book.lines)

#Each line is a character vector with a length of 1
book.lines[7]
length(book.lines[7])

#We can use the length function to inspect how many lines of text are in the file
length(book.lines)

```

#### **Quick R exercise - working with strings from non-tabulated data**{-}

1. Load the stringer package if you haven't yet

2. Use the str_which function to find all lines with the word title in them (remember to account for case sensitivity)

3. What is the title of the book you imported to R?

### **Exporting tabulated data**{-}

There are many ways to export data from R. In many cases you would be required to export tabulated data, and base R has an 
array of functions to handle the task. A convenient one is `write.csv()` function (yes it is a private case of 
`write.table()` function) which will export tabulated data as a CSV file. The tidyverse version is `write_csv()`, which 
works similarly however don't confuse them. 
Let's look at an example, the code snippet bellow will generate a 100X100 data frame, which will get exported as a csv 
file.
The snippet demonstrates once more the order of opperation

```{r write.csv}


#First let's generate the row names
#The paste function will generate a vector of 100 consecutive elements, Site1 - Site 100
#Note that I am not using a loop for that, the vectorization in R saves on a lot of iterations

sites <- paste("Site", seq(1,100, by=1), sep = "")

#Similarly l

cols <- paste("species", seq(1,100, by = 1), sep = "")

#Now lets generate the numbers, here I am cheating and using the replicate function
#It is a function that acts as a for loop, we will see some more of these function when we discuss itereations
#The replicate function will create a list with 100 elements each is a vector of a 100 random numbers

da <- replicate(100, sample(1:10000,100,replace = T), simplify = F)

#Now lets turn is to a dataframe, Reduce binds all the vectors as columns (again cheating and usig a function that acts as a for loop)
da.df <- as.data.frame(Reduce(f=cbind, da),row.names = sites)
names(da.df) <- cols

#Now we are ready to export the dataframe as a csv file

write.csv(da.df, file = "practice3.csv", row.names = T)

```

>export xls xlsx files

### **Exporting non-tabulated data**{-}

R allows you to export non-tabulated files as well, the following snippet demonstrates how to export non-tabulated data as a txt file. In future labs you will export other file types as well. 

```{r export txt files}
#Let's write the first 12 lines of the book as text file
#Note that if you were writing into an existing file you could use the argument append = T and R would add lines insted of writing over the file.

write(book.lines[1:12], file = "book.txt", append = F)

```

### **Quick R exercise - data exportation**{-}

1. Generate 3 strings and house each of them in a vector

2. Add the 3 strings as new lines at the end of book.txt 

## **(Tidy) Data manipulation in R**{-}

R has some powerful tools allowing you to manipulate and modify data. We have already gone over some of them. For example, 
you have learned several ways to subset objects, you have learned that almost any data type in R can be coerced to any 
other using the as.x functions. You have also learned the `rbind()` and `cbind()` can bind vectors to matrices and matrices
to larger matrices.

This section of the lab will expand your capabilities of modifying data, and will focus of tools mainly from the two 
Tidyverse packaged, dplyr and tidyr. This packages are designed to streamline data manipulation, making data more tidy. 
They share some of the philosophy and framework of [SQL](https://en.wikipedia.org/wiki/SQL), and thus are worth getting to 
know.

As always, there are some great cheatsheets for data manipulation using Tidyverse packages:

<details>
  <summary>Click to expand</summary>

* A Simple [cheetsheet](https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)

* A [comprehensive cheetsheet](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf)

</details>

***

### **A tidy re-imagining of the data frame**{-}

As mentioned, the tidyverse is an assortment of packages designed to tidy and streamline data analysis and code writing in 
R. As you will see the tidyverse redesigned several basic consents in R, including data structure such as factors (via the 
`forcats` package) and data frames via the tibble package. 
Tibbles are the default format of tabulated data for many tidy data importation packages.
What are tibbles? Tibbles are first and foremost data frames. Lets see an example using readr's `read_csv()` function.

```{r read_csv, message=FALSE,warning=FALSE}
df1_tbl <- readr::read_csv(file ="lab2_files/practice2.csv")
class(df1_tbl)
head(df1_tbl)
```
As you can see, tibbles print pretty similar to data frames and belong to the same class, which brings the following 
question.

#### **Why use tibbles?**{-}

There are several advantages to using tibbles, which are linked to the differences between them and traditional data 
frames. 
Tibbles are designed to handle larger data sets, which is why when we call them (lets look at that code again) they print 
differently.

```{r tib vs df1}
df1
df1_tbl

```
Note how when calling a tibble, it specifies its dimensions and the class of each column. if we were dealing with a larger 
data set the tibble would only print the first 10 rows and the columns to not overload the console. If you noticed another 
difference, don't worry we will deal with it soon.
Tibbles work great with the glimpse function that help inspect the data.

```{r glimpse}
glimpse(df1_tbl)
```
Tibbles are stricter than data frames in the sense that tibbles do not convert data types when created, for example you don't need to worry about it converting your character columns to factors.

Another fundamental difference between tibbles and data frames is that tibbles can be subset via piping....
Oh wait you don't know about those yet...

#### **Even the tidyverse can get messy sometimes**{-}

A guiding principle of tidy data is that every line of tabulated information should have a unique identifier (similar to 
SQL). In base R the row names act as a unique identifier for each row, as every row has a unique value. However, base R's 
approach allows you to store metadata (descriptive information about your data) in your row names as well. 
For example, the row name from `df1` you generated by importing the file practice2.csv has site names as row names. 
Tidyveres shares the same philosophy as SQLl and it does not view row names as a place to store descriptive data. 
This is not a big deal, it just requires that you assign the row names to a column.


```{r the tibble rn, eval=FALSE}
df1_tbl <- as_tibble(df1, rownames = NA)
df1_tbl
```

Not that it is still possible to keep the row names when converting to tibble

```{r the tibble rn2, echo=FALSE}
df1_tbl <- as_tibble(df1, rownames = NA)
DT::datatable(df1_tbl)
```



The code snippet bellow will introduce you to a tidyverse function that will allow you copy the row names as a 
column, I will also compare it to a non-tidyverse approach

```{r tidyvers vs non-tydiverse}

#Let's use the dataframe da.df as an example dataset

#First the non-tidyverse way, note that I am copying da.df sine I am planing to reuse it

#Usinf the cbind.dataframe function to column bind the row names and the and da.df to one dataframe

da.df2 <- cbind.data.frame(row.names(da.df),da.df)

#Change the name of the row names to Sites

names(da.df2)[1] <- "Sites"

#Inspect
da.df2[1:4,1:4]

#The tidyverse way

da.df3 <- tibble::rownames_to_column(da.df, "Sites")

da.df3[1:4,1:4]


```

### **Subseting columns and rows using base R and tidyverse**{-}

The dplyr package makes subseting columns and rows very simple using the select, filter, and slice functions.
The select function allows to subset columns, while the filter and slice functions allows you to subset rows function

```{r select }

#The select function allows you to subset by column

#Subset a range of columns

da.df3 <- select(da.df2, 45:55)

#Note that select does not remove the row names

da.df3[1:4,1:4]

#Alternatively it allows you to drop columns
#Keeping only the first 2 column

da.df3 <- select(da.df2, -(3:ncol(da.df2)))

da.df3[1:4,1:2]

#Select allows to choose specific columns

da.df3 <- select(da.df2, "species1", "species2","species11")

da.df3[1:4,1:3]

#From a list
species <- c("Sites","species13","species37","species55")

da.df3 <- select(da.df2, which(names(da.df2) %in% species))

da.df3[1:4,1:4]

#Lastely select can be used to reorgnize a dataframe
#Let's look at creating a column from row names again

da.df3 <- da.df

#Creating a column from row names, naming it sites
da.df3$Sites <- row.names(da.df3)
#Now sites is the last column
da.df3[1:4,ncol(da.df3)]
#, let's move it to be the first
da.df3 <- select(da.df3, Sites,everything())
da.df3[1:4,1:4]

```

***

#### **Quick R exercise**{-}

Select can be used with some helper functions from the tidyer package such as starts_with

1. Use select and a helper function to subset da.df for columns ending with 1 (hint, use on of the cheatsheets)

2. Use the select and the helper function one_of to select columns found in the species vector

***

Next lets see how to use filter and slice to subset a data frame by rows

```{r filter and slice}

#filter allows to subset the data frame based on values of one column
#Lets filter df based on CO2 concentration

df1

df2 <- filter(df1, CO2 > 300)

#Note that filter discards row name
df2

#slice allows to subset a range of rows

df2 <- slice(df1, 1:4)

#Note that slice discards row names as well
df2

```



### **Streamlining data manipulation using %>% (pip)operator**{-}

So far we have gone through two ways of sequentially running through functions.

1. Line by line

2. Chaining functions using brackets

X <- f3(f2(f1(x))) where the order in which the function run is inside out.

This sort of writing, while concise, makes the code hard to read. The Tidyverse employs a piping operator %>%, colloquially
referred to as an "and then" operator

x `%>%` f1() is equivalent to f1(x) and x `%>%` f1(y) is the equivalent of f1(x,y).

```{r piping}

#these lines are equivalent

da.df3 <- select(da.df2, 11:15)

da.df3 <- da.df2 %>% select(11:15)

#bonus subseting with a pipe, note that the dot (.) is a place holder for 
#the data frame at the beging of the pipe
df.df2.s13 <- da.df2 %>% .$species13 

```

As bonus 

The piping operator can be used to chain functions sequentially, from left to right allowing for a more readable code.

```{r}

#chaining functions, functions run from left to right on da.df2
#I am cheating here since you don't know what modify is, 
#though it is pretty self explanatory.
da.df3 <- da.df2 %>% select(1, 5:13) %>% filter(species11 > 50) %>% 
  modify_if(is.numeric, log1p)

```

***

>**A note about piping:**
> The `%>%` operator is part of a series of pipes [magrittr package](https://data-and-the-world.onrender.com/posts/magrittr-pipes/)
>Base R offeres pipes as well `|>`, however it is slighly less useable than `%>%`.
>I still encourage you to try using it. 


***


### **Adding and modifying columns**{-}

The dplyr package has functions dedicated to modifying and adding columns to tabulated objects.
The function mutate from the package dplyr can be used to add columns to data frames, as well as to modify them.  

```{r mutate1}

#Lets add a column called elevation, with 3 vauleus high, sea level, low to df
#Note that again I am copying it since I would like to use it again
# I am also creating the site collumn and than moving both to be columns 1 and 2

elevation <- c("High", "Sea level", "Low")

df2 <- df1 %>% mutate(Sites = row.names(df1)) %>% 
  mutate(Elevation = sample(elevation, nrow(df1), replace = T)) %>%
  select(Sites, Elevation, everything())

df2

#Mutate can also be used to modefy columns
#Lets log transform the CO2 column
#Again I am copying the dataframe since I plan on using it later

df3 <- df2 %>% mutate(CO2 = log1p(df2$CO2))

df2$CO2
df3$CO2


```

mutate has a subset of functions the modify all or a subset of columns from a dataframe, negating the need to use a loop.

```{r mutate 2}

#One useful mutate function is mutate_if, it allows to modefy a column if a condition is met
#For example it allows to change all character columns to factors

class(df3$Elevation)
df3 <- df2 %>% mutate_if(is.character, as.factor)
class(df3$Elevation)

#mutate_if allows to perform mathematical operations only on numeric columns
#e.g. log transformation and then multeply by 2 
#Note that ~ is used to escape the mathematical operator and the . is used to tell R to run the operation on the dataframe

df3 <- df2 %>% mutate_if(is.numeric, log1p) %>% 
  mutate_if(is.numeric, ~.*2)


``` 


### **Transforming dataframes**{-}

The dplyr and tidyr packages have many functions that allow to transform and reshape data frames.
Much like base R, dplyr's bind_rows and bind_cols pastes two data frames based on columns or rows. However, dplyr has some functions that allows for more sophisticated joining of two data frames

```{r joins}

#Let's use df2 and da.df2 as example, one contains information about species 
#and one contains information about chemical attributes. 
#However, df only has information about 10 sites, that's 90 less than in da.df2. 
#How can they be joined? 

#The function left_join takes two data frames as arguments and joins matching 
#values form tthe right to the left based on a joint column

#In this example I will select the first 5 columns of da.df2 
#(easier to see the results of the join) and join df to it based on the site column which is shared between the two.

da.df3 <- da.df2 %>% select(1:5)%>%
  mutate(Sites = str_to_lower(Sites)) %>% 
  left_join(df2, by="Sites")

#Note the NAs in rows that are not matched
da.df3 [10:12,]

#inner_join will keep only matching rows

da.df3 <- da.df2 %>% select(1:5)%>%
  mutate(Sites = str_to_lower(Sites)) %>% 
  inner_join(df2, by="Sites")

nrow(da.df3)

```

**Quick R exercise**

1. Try running left_join with df as the left argument, what was the result?

2. Try using right_join and full_join (you can use the dplyr cheetsheet for help)

***

So far the data used in this lab was stored in a wide format table, however, as you will see later in lab 2 there are 
advantages to using [long format tables](https://en.wikipedia.org/wiki/Wide_and_narrow_data). Long tables are ones where 
the columns are converted into keys and the content of the column is the value paired to each key. 

The gather function from tidyr package converts wide tables to long tables. 

```{r gather and split}

#Here using the gather function to convert df2 to wide table the column names are the key, the values they store are the values.
#Note that I am exampting Site and Elevation from it.

df3 <- df2 %>% gather(key = "Parameters", value = "Concentration", 3:ncol(df2))

df3

#The function spread converts long tables to wide tables
df3 <- df3 %>% spread(Parameters,Concentration)

head(df3)

```

***

dplyr can group data based on category and summarize it based on grouping.
This can be useful for example to calculate the average genetic diversity across geographical areas. 

```{r group and summrize}

#group_by is used to group the table by elevation, summarise is used to calculate the mean concntration of each chemical, per elevation. Because the Sites column is not numeric I used summerize_if.

df3 <- df2 %>% group_by(Elevation) %>% summarise_if(is.numeric, mean)

df3

```

#### **Quick R exercise**{-}

the summarize function can summarize more than just mean, and can take more than one argument to summarize by.

* Use list() within summarize_if to calculate both the mean, SD, and maximum value for each of the chemicals in df2. Assign in to df3

***

## **Conditional statements and iterations**{-}

Sometimes it is necessary to ask R to decide what type of code runs on which part of the data, or how many times to run each line of code. Conditional statement such as if else, and loops are there to do just that.

### **Conditional statements**{-}

Conditional statements are used to determine which code will handle what part of the data.
You have already used a conditional statement, even if you haven't noticed.
mutate_if is a function using a conditional statement modifying a column if a condition is met (e.g. is.numeric).

The basic conditional statement is if, it asks if a condition is met (e.g. if(x>8) = TRUE). Once the condition is met, something happens. if can be paired with an else statement to determine what to do if the condition is not met.

```{r if else}
# Here is an example of an if else statment note the curly brackets
x <- 8

if (x == 8) {
  print("x = 8")
} else { 
  print("x != 8")
  
}

#If else statments can be chaind

if(x>8) {
  print("x>8")
} else {
  if (x == 8) {
    print("x=8")
  } else {
    print("x<8")
  }
}


```

if else statements only work with scalars, i.e. only evaluate objects with one element.
In order to evaluate a vector, the function `ifelse()` should be used. `ifelse()` takes 3 arguments,
test (e.g. x>8), true (what to do in case test is true), false (what to do in case the test condition isn't met).

The following code snippet demonstrates the use of the `ifelse()` function and it's tidyverse equivalent `if_else()`.

```{r ifelse if_else}

#Creating a vector
x <- sample(1:1000, 10, replace = T)

#ifelse function, if x > 500 paste something else paste somthing else to y
y <- ifelse(x > 500, paste("x>500"), paste("x < = 500"))

#dplyr has a modefied ifelse function, it alows to handle missing values
z <- if_else(x > 500, paste("x>500"), paste("x < = 500"))

y
z

```

If else can also be applied to data frames as can be seen in the snippet bellow.
Note, that the code uses the

```{r if_else df}

#If elevation is high temp is -7 else temp is 10

df2 <- df2 %>% mutate(Temp = if_else(Elevation == "High", -7, 10))
```


In the previous 

```{r multiple ifelse}
#ifelse can be chaind, either as sequenctial funcation or within the first function.
#NAis a filler in case one condition is not met 
df2 <- df2 %>% mutate(Temp = if_else(Elevation == "High", -7, NA)) %>%
  #If test is false keep current value
  mutate(Temp = if_else(Elevation == "Sea level", 15, Temp)) %>%
  mutate(Temp = if_else(Elevation == "Low", 25, Temp))

df2

#alternatively
df2 <- df2 %>% select(-Temp)

#Note in second if_else, if_else is chaind in the false argument.
#This will not work with the first one since mutate needs to create the Temp column first
df2 <- df2 %>% mutate(Temp = if_else(Elevation == "High", -7, NA)) %>%
  mutate(Temp = if_else(Elevation == "Sea level", 15, 
                        if_else(Elevation == "Low", 25, Temp))) 

df2


```




Chaining multiple if else statements can make the code very unreadable and clumsy.
The `case_when()` function from the dplyr package acts as a multiple if else statements and accepts multiple test arguments.

```{r case_when, eval=FALSE}
#Let's creat a new column called contamination_level and assign values based on NH4 concentrations


df2 <- df2 %>% mutate(Contamination_level = case_when( NH4 > 8000 ~"Very high",
                                                       NH4 > 7000 ~"High",
                                                       NH4 > 6500 ~"Medium",
                                                       NH4 < 6500 ~"Low"))
df2


```

```{r case_when2, echo=FALSE,message=FALSE}
df2 <- df2 %>% mutate(Contamination_level = case_when( NH4 > 8000 ~"Very high",
                                                       NH4 > 7000 ~"High",
                                                       NH4 > 6500 ~"Medium",
                                                       NH4 < 6500 ~"Low"))
DT::datatable(df2)



```



***

#### **Quick R exercise - data grouping and summerizing 2**{-}

1. Group df2 by contamination level and summarize average chemical concentration per contamination level.

***

### **Iterations**{-}

As mentioned, R is pretty good at vectorized functions which negate the need for iterations.
For example ifelse negates the need to use a loop to iterate over each position of the inspected vector. Furthermore, we have worked with functions such as summarize that iterate over an entire dataframe or replicate which can be used to to generate a list of multiple elements.

These functions were made because loops are considered memory inefficient and slow, while vectorized functions and specialized iteration functions are compiled and efficient.

However, sometimes there is no escaping the loop. The only loop we will go over in this lab is the for loop, while there are other loops in R in my experience they are rarely used.

The for loop has the following basic structure for(element in vector) { do something}.
Normally a counting variable is used to subset elements while iterating over a vector
If a counting variable is used than the vector is a range of numbers.

Here is an example of a basic for loop

```{r basic for loop}

#Let's write a loop that picks lotomax numbers for you

#First assign a counting variable

i=1

#Generating random numbers
x <- sample(1:50, 8, replace = F)

#Constructing the loop
#seq_along() generates a vector of 1:length of x
#The reason I don't use 1:length(x) is that if the length of x is 0 the loop would still run, seq_along prevents this

for(i in seq_along(x)){
  print(paste("The",i, "number is:", x[i], sep = " "))
}

#Let's combine a conditional statment with the for loop and make a fancier number picker

for(i in seq_along(x)){
  if(1<= i & i <8){
  print(paste("The",i, "number is:", x[i], sep = " "))
  } else{ 
    print(paste("The bonus number is:", x[i], sep = " "))
  }
}



```

#### **Quick R exercise - pseudorandom numbers and loops**{-}

What would happen if set.seed(123) was the first line of the loop?

***

The previous snippet was a pretty simple use of a for loop, now lets look at a practical way of implement it. When working as a bioinformatician you will most likely have to load and manipulate data from multiple files. In the following example I will use read.csv to import all csv files in a certain folder and place them in a list. Note that I am generating the list outside the loop, when you index an existing vector rather than add new elements to it for loops run faster.

```{r practical for loop}

#Create a vector with a list of all csv files in my foulder.
#You can use the argument path = to change foulder. 
#recursive = F ignors subfolders
files <- dir(pattern = ".csv", recursive = F)

lsfile <- vector(mode = "list", length = length(files))

i = 1

for (i in seq_along(files)) {
  lsfile[[i]] <- read.csv(file = files[i],row.names = 1, stringsAsFactors = F)
  names(lsfile)[[i]] <- paste(files[i])
  
}  
  
```

***

## **Functions**{-}

If you remember, R lets users write customized functions.

As a bioinformatician you would probably be required to share scripts with other people. Sharing
customized functions instead of a messy script increases the reproducibility of your script because
all the receiving party has to do is input arguments instead of tinkering with endless lines of code.
Similarly, when you find yourself running the same scripts over and over it is advantageous to wrap
them up as customized functions. This clears up your script, and allows you to chain functions and
automate your analysis. This is what's known as a pipeline. 

As a student it also pays to know how write functions in R. If you find yourself conducting the
same analysis in different labs (say chemistry and physics), write a pipeline that generates the
graphs and tables you need. 

### **Constructing a function**{-}

The function structure is function(arguments){operations done on functions}. 
The arguments require input from the user, but they can have default values e.g.
function(arg1=5, arg2 = NA, arg3) { do something}. In this example arg1 and arg2 have default
values. They can be changed by the user, but if there is no input from the user the function will
use the default value. If the function is intended to return a value 

Lets look at two functions as an example, the first one is very simple the second a bit more complex. 

```{r function1}

#How to construct a function

#Assign a function to a variable, set the arguments
#Decide what the function does whith the arguments

print_x <- function(x){
  print("All this function does is print x")
  print(x)
}


# Let's turn the for loop lotto number generator to a functions


Lottery_nums <- function(npicks){
  #Need round numbers
  as.integer(npicks)
  #If 8 picks lotomax else lotto 649
  if(npicks == 8){
    print("Here are your lotomax numbers:")
    i=1
    nums <- sample(1:50, npicks, replace = F)
    for (i in seq_along(nums) ){
      if (i >= 1 & i <= npicks){
        print(paste("The",i, "number is:", nums[i], sep = " "))
      } else {print(paste("The bonus number is:", nums[i], sep = " "))
              }
    }
  } else {
    if( npicks == 7){
      print("Here are your lotto649 numbers")
      
      i=1
      nums <- sample(1:49, npicks, replace = F)
      for(i in seq_along(nums)){
        if(i >=1 & i <= npicks){
          print(paste("The",i, "number is:", nums[i], sep = " "))
        } else {print(paste("The bonus number is:", nums[i], sep = " "))
              }
      }
      
    } else{
      print("npicks has to be either 7 or 8")
      
      }
    
  }
 #Print disclamer 
  print("This number picker does not increase the chances of winning nor does it encourage participating in lottery games")
  
}

Lottery_nums(7)

```

***

Now that we have played around with some useless functions, lets write a practical function.
As I mentioned, when you work as a bioinformatician you would be required to perform the same task multiple times on multiple files and functions are great to automate that in a reproducible manner. 

In this example we will write a function that imports csv files,log transforms them and store them in a list. This is a very straight forward function, and its capabilities can be expended to perform other tasks.


```{r practical function}
#Note that I am giving each argument a default value

imp.data <- function(Pattern= ".csv", Path = getwd(), Recursive = FALSE){
  
  files <- dir(path = Path, recursive = Recursive, pattern = Pattern)
  i = 1
  f_list <- vector(mode = "list", length = length(files))
  
  for (i in seq_along(files)) {
    f_list[[i]] <- log10(read.csv(file = files[i], row.names = 1, stringsAsFactors = F) )
    
  }
  names(f_list) <- files
  
  #Note return
  return(f_list)
  
  
}





```

***

> **A note on the function environment**
>
>The function environment is distinct from the global environment, therefore if there is an interest in keeping some of the
>intermediate variables created by the function they need to be exported using the <<- operator. The <<- operator is
>variable assignment operator similar to <- only it exports variables to the global environment.

***

## **Plotting your data**{-}

R has a basic function to graph data, this is done with the plot function.
Here is an example of a scatter plot using the plot function.


```{r baseR plot}

plot(x=df2$NH4,y=df2$CO2, xlab = "NH4", ylab = "CO2", main = "A scatter plot")

```

As you can see the resulting graph is...OK. Luckily the tidyverse has a package that makes plotting graphs in R a whole lot better. 


### **Potting data with the ggplot2 package**{-}

The ggplot2 is a part of the Tidyverse packages. It's strength is that it allows to build graphs layer by layer based on the [grammer of graphics](https://vita.had.co.nz/papers/layered-grammar.pdf).

All plots generated by ggplot2 require the same three components:

* A data set
* A list of coordinates
* [geoms](https://ggplot2.tidyverse.org/reference/) - a set of marks that determines the geometry of the plot (e.g. box-plot, scatter plot)

As always ggplot2 has a great [cheetsheet](https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf).


#### **Basic ggplot2 plot**{-}

The plotting function in the ggplot2 package is ggplot, every plot starts with it.
In it you specify the data set the plot will use as well as the x and y coordinates.
On top of the basic information you add geoms, i.e. what graph element to add.
Elements of the graphs are added using the + sign.

Here is an example of a basic box plot


```{r ggplot1}

#ggplot works best with long tables, lets first convert df2 and log transform so that the diferences in values between chaicals will not be huge

df3l <- df2 %>% gather(key = Parameters, value = Concentration, 3:6) %>%
  mutate(Concentration = log10(Concentration))

#Lets generate the box plot.
#First start with ggplot and specify which data to use.
#The axis are specefied under the aes argument.
plot1 <- ggplot(data = df3l, aes(x=Parameters, y=Concentration))+
  geom_boxplot()

plot1

```

This is a nice plot, but it is possible to add more information to it.
For example overlay each observation (i.e. observation from each site) on the box plot using the geom_jitter function


```{r ggplot2}

#Let's overlay each observation on top of the box plot, this will allow you to see where the
#observation fall, and how many observations are in each category.
#On top of it, we will color each observation by the level of contamination in the site 
#from which it came

plot2 <- ggplot(data = df3l, aes(x=Parameters, y=Concentration))+
  geom_boxplot()+
  geom_jitter(aes(color = Contamination_level))

plot2


```

That is a nice graph, now lets see how to split a graph to panels by category (e.g. elevation).

```{r ggplot3}

#Facet_wrap allows you to split a graph into facets
plot3 <- ggplot(data = df3l, aes(x=Parameters, y=Concentration))+
  geom_boxplot()+
  geom_jitter(aes(color = Contamination_level))+
  facet_wrap(~Elevation)

plot3


```


Lastly, let's take care of the cosmetics. ggplot has a set of built-in themes, the default is classic view. However it's possible to change them using them_x, and further customize them using the theme function. The plot and axis title can also be modified.


```{r ggplot4}

# here I used theme_bw to switch to a black and white theme the base size argument changes font
#size
#I then used the theme function to further costumize the polt. Strip.text are the title of each facrt and axis.title are the title of the axis
#Basically every aspect of the graph can be modefied using theme or other functions

plot4 <- ggplot(data = df3l, aes(x=Parameters, y=Concentration))+
  geom_boxplot()+
  geom_jitter(size=2,aes(color=Contamination_level))+
  facet_wrap(~Elevation)+
  theme_bw(base_size = 16)+
  theme(strip.text = element_text(size = 16, face = "bold"),
        axis.title = element_text(face = "bold"))+
  xlab("Chemical parameters")+
  ylab("Concentration of chemicals")+
  ggtitle("A box plot of chemical parameters per elevation levels")
  

plot4


```


#### **Exporting plots**{-}

ggplot2 allows you to export plots using the ggsave function

```{r exporting plots, eval=FALSE}
ggsave(plot = plot4, filename = "a_plot.pdf", units = "in",
       width = 11.69, height = 8.27)
```


## **Summary exercise lab 2**{-}

Start a script file in RStudio name it 3315_lab1_yourinitials.R
You will submit this script and output file if applicable to D2l submission.

**1.** Load the mtcars data-set by using data(mtcars).

**2.** What class is mtcars? Show your code.

**3.** Once loaded assign it to the variable car_data, turn the row names into a column (name it Car_model), 
make sure it is the first column. Do this all in one line of code using the %>% operator and Tydiverse functions. Perform all following tasks on this variable.

**4.** The column am represents the car's transmission, 0=Automatic and 1 = Manual write a conditional statement that converts 0 and 1 to Automatic and Manual respectively. Do it within a mutate function. Move am to be the 2nd column of the dataframe. Do it in a single line of code

**5.** Remove the columns vs, gear and carb from car_data using a Tydiverse function

**6.** Use ggplot to plot a box plot of mpg, with cyl as the grouping category (if you're getting an error thing what types of values are categorical values). Add jitter to the plot, make sure the size of the dots is 2. Make sure that the x axis title is Number of cylinders and the y axis title is Miles per gallon. Brownie points for any other cosmetic modification of the plot.

**7.** Export the plot as pdf using a ggplot function, attach the file to your submission.

**8.** Write a for loop that generates three dataframes, the dataframes needs to have 2 numerical columns and 1 character columns and 100 rows. 
The loop should store each dataframe as an element of a list and save each dataframe as a csv file. The names should be dataframe1.csv - dataframe3.csv. You can use the paste function with the loop counting variable to make sure the names are not hard coded. Hint, make the list before you loop.

**9.** Write a function that reads all csv files in a folder, log1p transforms all the numeric columns and stores the data as an element of a list. Hint, make the list before you loop and make sure to return the list as the final step of the function.
