```{r setup lab3}
#| include: false
knitr::opts_chunk$set(echo = TRUE, comment = "", prompt = TRUE ,class.source = "fold-show")
options(knitr.duplicate.label = "allow")
```

```{r set libs}
#| include: false
#| message: false
#| warning: false
lapply(c("vegan","ape","ade4","ggdendro","dendextend","seqinr","pvclust", 
         "tidyverse", "Biostrings"), 
       library, character.only = TRUE)
```

# **Lab 3: Handling of Biological Data in R - 1**

<br>

## **Learning objectives** 

<details>
  <summary>Click to expand</summary>
__By the end of this lab you will be able to:__

* Demonstrate the ability to install packages from Bioconductor

*	Demonstrate the ability to import and export biological sequences

*	Demonstrate the ability to manipulate biological sequences using base R and tidyverse

* Demonstrate the ability to collect basic statistics of biological sequences using base R

* Demonstrate an understanding of the basic data classes of the Biostrings and seqinr packages

* Demonstrate the ability to manipulate both multiple and single biological strings using tools from the Biostrings and seqinr packages

* Demonstrate the ability to collect basic statistics of biological strings using tools from the Biostrings and seqinr packages

</details>

***

# **Lab3**

R has many tools for handling biological data. Over the next couple of labs, you will build the vocabulary needed to work with sequences, statistical methods and objects, trees, and dendrograms.

```{r install packages}
#| eval: false
#| echo: true
install.packages(c("vegan","ape","ade4","ggdendro","dendextend","seqinr","pvclust"))

# lapply is part of the apply family of functions, one of those functions that act as a for loop
lapply(c("vegan","ape","ade4","ggdendro","dendextend","seqinr","pvclust"), 
       library, character.only = TRUE)
```

You should also load the tidyverse packages.

```{r load tidyverse}
#| eval: false
#| echo: true
library(tidyverse)
```

## **Installing packages from Bioconductor**

[Bioconductor](https://bioconductor.org/) is a repository that stores R packages geared towards the analysis of biological data. Most packages in Bioconductor are designed to handle omics, cellular, genetic, and biochemical data, and thus will be more useful in more advanced courses. However, some of the packages on Bioconductor are excellent for this course. One of them is the Biostrings package, which, as its name suggests, handles biological strings (i.e., nucleotide and amino acid sequences). While there are packages on CRAN with similar functions, it is a good opportunity to familiarize yourself with the object structure employed by Bioconductor packages.

```{r installing bioconductor}
#| eval: false
#| echo: true
# Bioconductor requires its own package manager
install.packages("BiocManager")

# Once the manager is installed, you can install Bioconductor packages
BiocManager::install("Biostrings")

library(Biostrings)
```

Note that when you load the packages, functions can get masked. Make sure you are using the function from the correct package. You can do that by specifying the package when calling the function, e.g., `dplyr::summarise()`.

## **Working with sequences using base R and the tidyverse**

You were introduced to strings during lab 1, and biological strings are no different than regular strings. Nucleotide sequences are comprised of up to 20 nucleotide symbols (ATCG, U, and letters that denote alternative bases, base groups, and modified bases). Amino acid sequences are comprised of up to 28 symbols (the 20 standard amino acids plus special codes for ambiguity and modifications).

Base R and some functions from `stringr` make it easy to perform basic processing of biological strings. Let's reexamine a simple example and create a string of 100 nucleotides.

```{r generate sequence}
# Why am I setting seeds?
set.seed(649)
# Generate a vector of nucleotides
sq1 <- paste(sample(c("A","T","C","G"), 100, replace = T), collapse = "")
```

Next, we can count how many times each nucleotide appears in the sequence using a simple `for` loop.

```{r count nucleotides}
nucs <- c("A","T","C","G")
reps <- rep(0, 4)

for (i in seq_along(nucs)) {
  reps[i] <- str_count(sq1, nucs[i])
}

df <- data.frame(nucs, reps)
df
```

It is also simple to chart these counts.

```{r plot sequence counts}
ggplot(df, aes(x = nucs, y = reps)) +
  geom_bar(stat = "identity")
```

It is relatively simple to find start and stop codons (the first step in predicting ORFs).

```{r find codons}
# With the stringr package it is simple to find all start and stop codons
(atg <- str_locate_all(sq1, "ATG"))

# Just one of the stop codons
(stp <- str_locate_all(sq1, "TAA"))
```

From here you can write a loop calculating the length of characters from the first nucleotide of each start codon to the last nucleotide of every stop codon. Those that are divisible by 3 represent open reading frames. We won't implement this here. The `stringr` package also makes it easy to convert between DNA and RNA.

```{r convert to rna}
# stringr also makes it easy to convert between DNA and RNA
sq1.rna <- str_replace_all(sq1, "T", "U")

str_sub(sq1, 1, 10)
str_sub(sq1.rna, 1, 10)
```

***

**Quick R exercise**

1. Write a function that takes a nucleotide string, summarizes the occurrence of each nucleotide into a table, and graphs it.

***

Next, let's see how to convert a sequence of nucleotides to its reverse and complementary sequence.

```{r sequence replace one base}
# First let's generate a shorter sequence, it will be easier to inspect
sq2 <- str_sub(sq1, 1, 10)

# stringr doesn't have a good option to replace every base with its complement because it can only 
# replace one pattern at a time.
sq2c <- str_replace_all(sq2, "T", "A")
```

Now that all Ts are As, there's no way to replace all original As with Ts. It can be done by splitting the string into a character vector using `str_split` from `stringr`. Note the `[[]]` since `str_split` outputs a list.

```{r complement chain}
sq2c <- str_split(sq2, "")[[1]]

# Now with a chain of ifelse it is possible to replace each position
sq2c <- if_else(sq2c == "T", "A",
                if_else(sq2c == "A", "T", 
                        if_else(sq2c == "G", "C", if_else(sq2c == "C", "G", sq2c))))

# Complementary, indeed.
sq2
sq2c
```

We can now use the `rev` function to reverse the order of the vector, resulting in a sequence that is reverse and complementary.

```{r reverse vector}
sq2
sq2c
(sq2rc <- rev(sq2c))

# Is it possible to locate a codon using a vector?
# First occurrence using str_locate
str_locate(sq2c, "GTT")[1,]
```

In order to find the codon, we need to find all occurrences of the nucleotide G, and then find which nucleotides are found one and two positions over. Not impossible but definitely annoying. An easier way would be to convert the vector back to a string and match patterns.

***

**Quick R exercise**

We will need to convert sequences to reverse complement again. Based on the code snippet above, write a function that does so. Name it rev_comp.

```{r rev_comp}
#| echo: false
rev_comp <- function(x){
  
  if(length(x) == 1){
    x <- str_split(x, pattern = "")[[1]]
  }
  x_rc <- x
  x_rc <- if_else(x_rc == "T", "A",
                if_else(x_rc == "A", "T", 
                        if_else(x_rc == "G", "C", if_else(x_rc == "C", "G", x_rc))))
  x_rc <- paste(rev(x_rc), collapse = "")
  return(x_rc)
  
}
```

***

Converting a string to its reverse and complement sequence adds a level of complexity because of the switch between strings to character vectors and back. This doesn't look too complex, but as bioinformaticians you will never work with just one sequence. You will use FASTA files with multiple sequences. Let's see how base R handles those. 

**Importing and working with FASTA files**

First, what are FASTA files?

FASTA files are text files in the [FASTA format](https://en.wikipedia.org/wiki/FASTA_format). The FASTA format requires that the first line of each sequence be a header, denoted by `>` (greater than) followed by a description of the sequence, e.g., `>E.coli strain K12 complete chromosome`. Next are all the letters denoting nucleotides or amino acids with no spaces (basically a string). Sequences in FASTA format are universally used by all sequence repositories, and FASTA files (generally .fa or .fasta extensions, though any txt file with the format will work) are universally used. Because FASTA files are essentially text files, it is easy to import them using the `file()` and `readLines()` functions.

```{r reading fasta files}
# Let's import 
nuc_fa <- readLines(
  file(
    "https://raw.githubusercontent.com/idohatam/Biol-3315-files/main/practice_nuc.fa",
    open = "r"))
```

Now, let's inspect the file.

```{r fasta inspect}
# In this file each sequence is contained in multiple lines
# As can be seen in this example, lines 2-4 still belong to sequence 1 
nuc_fa[1:4]

# Let's use a shortcut to do that, and see that the file contains 5 sequences
(length(hdr <- str_which(nuc_fa, ">")))
```

Now that we know the beginning and the end line of each sequence, we can gather some stats. Let's see how long each sequence is using an elegant for loop.

```{r sequence length loop}
# Note: no hard coding. This way, you can reuse this code even if the next file has more sequences.
seq_length <- rep(0, length(hdr))

# The for loop
for (i in seq_along(hdr)) {
  # What is the role of this if else statement?
  if(i < length(hdr)){
    seq_length[i] <- sum(nchar(nuc_fa[(hdr[i]+1):(hdr[i+1]-1)]))
  } else { 
      seq_length[i] <- sum(nchar(nuc_fa[(hdr[i]+1):length(nuc_fa)]))
    }
}

# Let's get the stats on sequence length
summary(seq_length)
```

We can see that the longest sequence is 786 nucleotides, and the shortest is 544 nucleotides.

What if there is a need to perform more tasks with the sequences? For example, find ORFs, compile the reverse and complementary sequences, or align the sequences against each other, build a phylogenetic tree, etc.? 

The fact that the sequences are divided into lines (different elements of a character vector) makes it difficult to process them, and they need to be housed in a new container. Two possible ways to do it are:

* House them in a dataframe: column 1 will be the header, column 2 will be the concatenated line of each sequence, and column 3 will be the reverse complement sequence.

* House each of them in a list: the name of each element will be the header, and a second list would be the reverse complement sequence. The two lists could be merged as two nested lists. 

Both options will require a fair bit of code. Let's see what the dataframe option looks like.

```{r housing seqs in df}
#| echo: true
# Let's start with creating 3 vectors

# Sequence headers
hdrs <- nuc_fa[hdr]

# A vector to house the seqs
sqs <- rep("H", length(hdr))

# Vector to house the reverse complement seqs
sqs_rc <- sqs
```

Now let's write a for loop to merge all lines of each sequence to one sequence and also produce the reverse complementary sequence.

```{r merge seqs df}
for(i in seq_along(hdr)){
  
  if(i < length(hdr)){
    
    sqs[i] <- paste(nuc_fa[(hdr[i]+1):(hdr[i+1]-1)], collapse = "")
    
    # My function also makes sure to check if the string is split or not and splits it if needed.
    sqs_rc[i] <- rev_comp(sqs[i])
    
  } else{
    
    sqs[i] <- paste(nuc_fa[(hdr[i]+1):length(nuc_fa)], collapse = "")
    
    sqs_rc[i] <- rev_comp(sqs[i])
    
  }
}

# Now let's bind the vectors as dataframe
df_sqs <- data.frame(hdrs, sqs, sqs_rc, stringsAsFactors = F)
```

Now that the sequences are housed in a dataframe, it is easy to get what we need from them.

```{r sequence calculations}
# Calculate
df_sqs <- df_sqs %>% mutate(Seq.len = nchar(sqs))

# Get sequence stats
summary(df_sqs$Seq.len)

# Get all ATGs
ATGs <- str_locate_all(df_sqs$sqs, "ATG")
# Result is a list with 5 elements, each with the positions of ATG of a particular sequence
# This is for the first sequence
ATGs[[1]]
```

Because dataframes require that all columns be of the same length, they are not very flexible and only allow you to work with strings. The second option, working with lists, affords more flexibility since each element of the list is independent of each other and can even be another list. Therefore, you could generate a set of nested lists where the outer list will house all other lists, and each nested list will contain the string and vectorized sequences.

Example:

List of all sequences (

  List of sequence 1 (element one (string of sequence), element two (string of reverse complement),
                      element 3 (character vector of sequence), element 4 (character vector of
                      reverse complement))
  .
  .
  .
  List of sequence 5 (....)
)

From there you can add more elements such as location of all ORFs, corresponding RNA and amino acid sequences. However, as you can see, all these will require **a lot of coding!** Luckily you don't have to reinvent the wheel since there are a few R packages designed to handle biological strings. This, however, was not a pointless exercise because you gained an understanding of some of the data structures generated by these packages, and hopefully it will make it easier for you to work with them. Furthermore, now you know what it takes to write a package like the ones you will use. This understanding will serve you when you will be required to develop your own packages when you work as a bioinformatician.

## **Biostrings**

The first package we'll work with is `Biostrings`. As its name suggests, this is a package that handles biological strings. It can handle DNA, RNA, and amino acid sequences and is also capable of pattern matching, which helps with sequence look-up and alignment (albeit limited in that respect). `Biostrings` allows you to import individual sequences or FASTA files containing multiple sequences.

Biostrings uses a specialized container class called `XString` to store sequences. `XString` is similar to a character vector; however, unlike a vector, it can only house a single string.

The `XString` class has four subclasses:

1. `DNAString` - stores DNA sequences

2. `RNAString` - stores RNA sequences

3. `AAString` - stores amino acid sequences

4. `BString` - stores general strings

Let's look at a few examples of working with strings using `Biostrings`.

```{r reading sequence biostrings}
#| echo: true
#| error: true
# As always, let's generate a random string and convert it to an XString object
string <- "This is a random string (is it random though?!)"

# Now let's call it as DNAString
string <- DNAString(string)
```

Uh-oh, an error message. This is because `DNAString` is limited in which characters it accepts. Remember, only `BString` is character-agnostic.

Let's see which characters `DNAString` accepts.

```{r xstring characters}
DNA_ALPHABET

# In case you were wondering what each of those letters mean, Biostrings has a function with the 
# details. Useful when designing synthetic sequences like primers

IUPAC_CODE_MAP
```

As you can see, `DNAString` accepts more than just ATCG. It also accepts base categories, strand orientation (+/-), and alignment gaps.

Let's see which characters the other two restricted subclasses accept.

```{r xstring characters-2}
RNA_ALPHABET

AA_ALPHABET
```

Now, let's use the acceptable alphabet to generate a string that will not produce an error.

```{r xstring}
# I am creating an AA sequence, using AA_STANDARD to just include the standard amino acid symbols
# Note that I am setting seed. If you want to reproduce the same random string I produced, 
# you should set seed to the same number

set.seed(123)

casein <- AAString(paste(sample(AA_STANDARD, 10, replace = T), collapse = ""))

casein

class(casein)

# Note, typeof will output S4. This is a different object type than ones we have worked with so far
typeof(casein)

str(casein)
```

One advantage of the `XString` class is that it allows you to treat the sequence both as a vector and as a string, making it easy to index, subset, and modify.

The code snippet below demonstrates how to access and modify an `XString` class object.

```{r xstring2}
# Let's get the first element of the AA sequence
casein[1]

# Let's subset
casein[1:5]

# Another great feature of the XString class is that the string is mutable in the same way a vector 
# is. Let's change the first AA to methionine.

casein[1] <- "M"

# Appending characters is done in the same way you would a character vector.
casein[11] <- "J"
# Reminder that gaps can also be added.
casein[10] <- "-"
```

Because XString class objects are treated as strings, it is easy to use functions from base R to process them. Let's use `grepl()`, which works like `str_detect` for pattern matching. Note that I am calling `base::` because `grepl` is masked by `BiocGenerics`, a Bioconductor package. It will work the same if you choose to use the `BiocGenerics` function.

```{r fu_tidyverse}
base::grepl("MYFNI", casein)
```

`tidyverse` packages are more particular about the data types they accept and require some manipulation. I used the base R function `toString()` to turn casein into... a string.

```{r futv2}
str_detect(toString(casein), "MYFNI")
```

It is also possible to use base R functions like `gsub()` and `stringr` functions like `str_replace()` to edit casein; however, it takes a bit of tinkering to keep it as an XString element. Still, it is worthwhile if editing chunks of sequence. Note that here I am also introducing you to base R's native pipe operator `|>`. We won't use it often, but it's worth knowing.

```{r native_pipe}
# tidyverse
casein2 <- casein %>% toString %>% str_replace("MYFNI", "MYXGT") %>% AAString()

# base R with the native pipe
casein3 <- gsub("MYFNI", "MYXGT", casein) |> AAString()
```

The Biostrings package has some useful functions with respect to sequence processing. Let's look at some of them.

```{r biostrings functions}
# Biostrings can sum up the frequency of each letter (i.e., AA in this case)
x <- alphabetFrequency(casein)
# Output is a named vector, not the most convenient way to present results.
x
# Can subset to only include those larger than 0 (and normalize to percent)
# as.prob gives the probability of each AA
x <- alphabetFrequency(casein, as.prob = T)[alphabetFrequency(casein) > 0] * 100

x
```

Let's do the same to a longer sequence, this time DNA. First, convert the `sq1` string to `XString`.

```{r biostrings frequency dna}
# First, convert the sq1 string to XString
sq1.bio <- DNAString(sq1)

# If using nucleotides, can use the argument baseOnly to only get ATCG
x.d <- alphabetFrequency(sq1.bio, baseOnly = T, as.prob = T) * 100

x.d

# Alternatively, it's possible to look at the proportion of only one base with letterFrequency
letterFrequency(sq1.bio, "C", as.prob = T) * 100
letterFrequency(sq1.bio, c("G", "C"), as.prob = T) * 100
```

Now let's convert the frequency information of all nucleotides to a dataframe and plot it.

```{r biostrings frequency plot}
x.d.df <- data.frame(names(x.d), x.d, stringsAsFactors = F)

ggplot(x.d.df, aes(y = x.d, x = names.x.d.)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  theme_bw(base_size = 16) +
  xlab("Bases") +
  ylab("Frequency")
```

Biostrings allows you to reverse, complement, and even translate nucleotide sequences.

```{r reversing}
# Reverse a sequence
sq1.bio.rev <- Biostrings::reverse(sq1.bio)

# Complement a sequence
sq1.bio.comp <- complement(sq1.bio)

# Reverse complement a sequence
sq1.bio.rev.comp <- reverseComplement(sq1.bio)

sq1.bio
sq1.bio.rev
sq1.bio.comp
sq1.bio.rev.comp

# Translate a sequence
sq1.bio.t <- translate(sq1.bio)

sq1.bio.t
```

***
**Quick R exercise**
Use `?Biostrings::translate` to see what arguments the function takes. Can it be used to predict ORFs?

***

**Working with multiple sequences**

As you know by now, bioinformaticians do not work with a single sequence but rather with many sequences in parallel. `Biostrings` offers the class `XStringSet` as a container for multiple sequences. Here too, it is a combination of a list and a vector. Think of it as a list of XString class objects.

Just like with the `XString`, the `XStringSet` offers four subclasses:

1. `DNAStringSet` to house DNA sequences

2. `RNAStringSet` to house RNA sequences

3. `AAStringSet` to house amino acid sequences

4. `BStringSet` to house sets of diverse strings

This means that a `BStringSet` can house the DNA, RNA, and AA sequences of a particular gene.

Let's see what it looks like.

```{r xstringset}
# Let's generate a DNAStringSet
sq1.set <- DNAStringSet(c(paste(sq1.bio), paste(sq1.bio.rev),
                          paste(sq1.bio.comp), paste(sq1.bio.rev.comp)))

# As you can see, an XStringSet object shows a glimpse of the sequence and the length of each 
# sequence
sq1.set

# Sequences housed in an XStringSet object can have names. Let's name our sequences
names(sq1.set) <- c("sq1.bio", "sq1.bio.rev", "sq1.bio.comp", "sq1.bio.rev.comp")

sq1.set
```

By now you probably realized that the `XStringSet` can house sequences from FASTA files. Let's import the FASTA file practice_nuc.fa as an `XStringSet` object.

```{r importing fasta as XStringSet}
# Using the readDNAStringSet function to import the FASTA file.
# Note that if the FASTA file would have contained both AA and nucleotides, you could have imported 
# it using the readBStringSet function
fa.set <- readDNAStringSet("https://raw.githubusercontent.com/idohatam/Biol-3315-files/main/practice_nuc.fa")

# As you can see, the imported file is organized very conveniently
fa.set
```

Objects of class `XStringSet` can be subset, indexed, and modified similar to objects of class `XString`.

```{r working with XStringSet}
# How many elements are in the XStringSet?
length(fa.set)

# What is the length of each of the sequences?
width(fa.set)

# Subset the first element
# Note that when subsetting an XStringSet, the output is an XStringSet and not an XString
fa.set[1]

# Because XStringSet objects are named, they can be indexed and subset using the $ operator.
# In this case the output is an XString object and it can be treated as such
fa.set$`k141_185580_GM_enrich_2.fa:145193-145801 `

# First character of the XString
fa.set$`k141_185580_GM_enrich_2.fa:145193-145801 `[1]

# Can be modified like a regular vector
fa.set$`k141_185580_GM_enrich_2.fa:145193-145801 `[1] <- "G"

fa.set$`k141_185580_GM_enrich_2.fa:145193-145801 `[1]

# Because XStringSet objects behave like a list, strings can be accessed using "[[ ]]"
fa.set[[1]][1] <- "A"

fa.set[[1]][1]

# It is easy to access a range of sequences
fa.set[1:3]
```

One of the advantages of an `XStringSet` is the ability to manipulate multiple sequences at the same time. 

```{r XStringSet working}
# alphabetFrequency can calculate base frequency for every sequence. Output is a matrix. Let's 
# output as dataframe
fa.set.frq <- as.data.frame(alphabetFrequency(fa.set, baseOnly = T, as.prob = T) * 100)

# Let's organize it a bit, add the names and plot
fa.set.frq <- fa.set.frq %>%  mutate(Sqs = names(fa.set)) %>% select(Sqs, everything(), -other) %>%
  gather(key = "Bases", value = "Frequency", 2:5)

# Not the prettiest, but you can see that the one at the top left has a much lower GC content than 
# the rest
ggplot(fa.set.frq, aes(x = Bases, y = Frequency)) +
  geom_bar(stat = "identity") +
  facet_wrap(~Sqs) +
  theme_bw()
```

Similarly, sequences can be reversed, complemented, reverse complemented, and translated in bulk.

```{r reversing2}
# Reverse complement
fa.set.rc <- reverseComplement(fa.set)

fa.set

fa.set.rc

# Pattern matching can also be done for all the sequences. Let's search for start codons
str_locate_all(toString(fa.set), "ATG")
```

## **seqinr**

An alternative package to `Biostrings` is `seqinr`. It is not a Bioconductor package and thus uses standard R objects. Much less elegant than `Biostrings`, yet it has some perks that are worth getting to know.

First, let's see how `seqinr` handles sequences by importing the practice_nuc FASTA file.

```{r seqinr import}
# Import the file
fa.snr <- read.fasta(file = "https://raw.githubusercontent.com/idohatam/Biol-3315-files/main/practice_nuc.fa")

# Note that this is a list
str(fa.snr)
```

As can be seen, the output from `read.fasta()` is a list. Each sequence is an element of the list stored as a character vector, resembling the exercise we did at the beginning of the class. Another thing to note is that the characters are not capitalized. This means that when pattern matching is required, the vectors need to be converted to strings. `seqinr` has two functions: `c2s()` and `s2c()`, which convert characters to strings and vice versa, respectively. 

Let's look at an example for pattern matching with `seqinr` by identifying all start codons in all the sequences in the file. 

```{r seqinr match patterns}
# Since fa.snr is a list, lapply is used to run c2s on all elements of the list
# Remember that the apply family of functions is basically an optimized for loop
# lapply takes a list and runs a function on all its elements and outputs a list with the results
# Then str_locate_all is used to find all ATG. (?i) is for case insensitivity
str_locate_all(lapply(fa.snr, c2s), "(?i)ATG")
```

Apart from concatenating vectors and splitting strings, `seqinr` has some summary functions as well.

```{r seqinr2}
#| echo: true
# count is a seqinr function that counts the occurrence of bases or a combination of bases
# wordsize is the argument that controls combinations. When set to 1 it counts the occurrence of
# each nucleotide. If it was set to 3 it would count the occurrence of each codon.
# Note that within lapply I am specifying which package to use count from. 
# This is because we loaded multiple packages with a function named count and I want to make sure 
# lapply uses the correct one

lapply(fa.snr, seqinr::count, wordsize = 1)
```

The function `GC` from `seqinr` calculates %GC. Just as `count`, it does not access a list, and you have to present it with a vector, so I used `lapply` to iterate over each of the sequences.

```{r seqinr2 gc}
lapply(fa.snr, GC)

# Let's get it into a plottable dataframe
# Note how I chain lapplys here. These are the equivalent of a nested for loop
fa.snr.gc <- as.data.frame(lapply(lapply(fa.snr, GC), cbind)) %>%
  gather(key = "Sqs", value = "GC") %>% mutate(GC = GC * 100)
```

***

**Quick R exercise**

Use `ggplot2` to plot a bar graph of the GC content we just calculated.

***

While `seqinr` does not seem impressive compared to `Biostrings`, it does have some perks. One of them is the ability to download sequences straight from releases of online databases such as GenBank and EMBL stored on servers of the University of Lyon in France. It's worth noting that it is not the only package that can do that. `ape` is another package that can do that, and you will be introduced to it in lab #4.

First, let's inspect which sequence repositories `seqinr` can interface with.

```{r repositories}
#| eval: false
#| echo: true
# Getting the list of repositories
# Note, to prevent masking, the code specifies to use choosebank from the seqinr package

repos <- seqinr::choosebank()
# Large list 
repos
```

In order to download sequences from a database, you must first select it using the `choosebank()` function. Here we will work with `genbank`, a comprehensive publicly available database of genetic sequences maintained by the NIH. 

```{r repositories 2}
#| eval: false
#| echo: true
# Chooses genbank
bank <- seqinr::choosebank(bank = "genbank", infobank = T)

# Let's inspect the database
bank
```

As you can see, unfortunately the university works with a release that isn't very up-to-date. Next, let's download some sequences. Since `genbank` is a massive repository with tens, if not hundreds, of GB of data, let's be targeted and download only sequences associated with the bacterium _Escherichia coli_.

To download the sequences, we will use the `query` function. The `listname` argument names your list, and the `query` argument tells the function which sequences to download. FYI 1: the `query` argument accepts regular expressions, which can simplify things. FYI 2: having an argument with an identical name to the function is not great practice. 

```{r repositories 3}
#| eval: false
#| echo: true
ecoliq <- seqinr::query(listname = "ecoli", query = "sp=Escherichia coli")
```

Now let's see how many sequences associated with _E. coli_ are in our database, and download the first 10 with the function `getSequence`. Note the `query` outputs the sequences as a list. That's why we need to use the function `unlist` to get a character vector. 

```{r repositories 4}
#| eval: false
#| echo: true
# See how many seqs
length(ecoliq$req)

ecoli10 <- unlist(getSequence(ecoliq$req[1:10], as.string = TRUE))

# Can treat individual elements as strings
# Get the first 10 characters of the first sequence.
substring(ecoli10[1], 1, 10)
```

As you can see, downloading sequences using seqinr stores the information in very basic containers. The next step will be to convert them to an `XStringSet` object. 

```{r seqinr to Biostrings}
#| eval: false
#| echo: true
# DNAStringSet accepts vectors with strings as elements. That's why we used unlist in the previous 
# step
ecoli10dss <- DNAStringSet(ecoli10)

# Let's add the sequences' names to the XStringSet object.
# This is a nifty trick using lapply in lieu of a for loop.
# The [[ part tells lapply to use square brackets to subset each element of the list, and the 1 
# tells it what position.
# Unlist turns the subset list into a vector that can then be used for the names function 

names(ecoli10dss) <- unlist(lapply(ecoliq$req[1:10], '[[', 1))
```

From here it's possible to calculate all the sequence stats you were introduced to at the beginning of the lab.

## **Exporting sequence elements**

Now that we got all of these sequences into R, how do we get them out? As always, there are several options.

Let's start with the easiest option: writing a file using Biostrings.

```{r biostrings export}
#| eval: false
#| echo: true
# writeXStringSet exports XStringSets as FASTA files 
# When you open the file you will notice that the name of each sequence is now the header 
writeXStringSet(ecoli10dss, "ecoli1.fa")
```

seqinr also allows you to export objects as FASTA files.

```{r write.fasta}
# write.fasta is a seqinr function to save objects as FASTA files
write.fasta(sequences = fa.snr, names = names(fa.snr), file.out = "practice2.fa")
```

Lastly, base R write function allows you to also write FASTA files (since they are basically text files). Not as elegant since it requires a loop, which really slows down the operation. 

```{r baseR write}
#| eval: false
#| echo: true
i = 1

for (i in seq_along(ecoli10dss)) {
  # This part writes the header
  write(paste(">", names(ecoli10dss[i]), sep = ""), file = "ecoli2.fa", append = T)
  # This part writes the sequence
  write(paste0(ecoli10dss[i]), file = "ecoli2.fa", append = T)
}
```

What do we do with the files now that we have them saved? You'll see in the upcoming labs.

# **Summary exercise:**

Start a script file in RStudio. Name it 3315_lab3_yourinitials.R. You will submit this script and output file (if applicable) to Brightspace.

**1.** Write a function that reads FASTA files and counts how many sequences are present in the files and how long each sequence is.

**2.** Write a function that takes a nucleotide sequence as a string and reverse complements it. Make sure that the function differentiates between DNA input and RNA input. Make sure the output is a DNA string.

**3.** Write a for loop that generates five 500 amino acid strings. Make sure to only use the standard 20 AA alphabet.

  + Make sure that the strings are <u>**NOT**</u> identical
  
  + The loop should store each string as an element of a vector named my_fave_proteins (hint: make the vector before the loop).
  
**4.** Assign the vector my_fave_proteins to an `XStringSet` object. Name that object my_fave_proteins_set.

**5.** Change the first AA of each protein in my_fave_proteins_set to M and give each protein a creative name.

**6.** Get the frequency of each AA in each protein stored in my_fave_proteins_set and plot a box plot of the frequency of each AA in all the proteins (plot using `ggplot2`).

  + Export the plot as PDF
  
**7.** Export my_fave_proteins_set as a FASTA file. Name the file yourfirstname_fave_prots.fasta.

**8.** Use `seqinr` to import the first 200 Archaea-associated sequences from `genbank`. How many Archaeal sequences are there in total? 

**9.** Assign the imported sequences to an `XStringSet` object named my_fave_arch_set and reverse complement the sequences.

  + Do not forget to include the sequences' names
  
**10.** What is the length of the longest sequence out of the 200 you imported? Show your work.

**11.** Export my_fave_arch_set as a FASTA file. Name the file yourfirstname_fave_arch.fa.
