```{r setup lab4, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "", prompt = TRUE ,class.source = "fold-show")
options(knitr.duplicate.label = "allow")

```


# **Lab 4 handling of biological data in R - 2**{-}

**Learning objectives** 

__By the end of this lab you will be able to:__


*	Demonstrate the ability to generate distance matrices from verity of data types

* Plot distances 

* Demonstrate the ability to generate dendrograms 

* Demonstrate the ability to generate phylogenetic trees

* Demonstrate the ability to export and import trees and dendrograms

* Demonstrate the ability to plot trees and dendrograms



## **The setup**{-}

In the previous lab you were introduced to handling biological strings with R. In this lab you 
will learn how to handle some of the output formats that result from analyzing those strings in R.
In most cases, as you will see in lecture and subsequent labs, this entails working with distance 
matrices and some sort of a hierarchical clustering of those matrices.

R's basic data structures, mainly lists, provide very flexible containers to house such complex type
of information. As you will see, most non-bioconductor packages house their outputs in list form or
at least as a list based object.

For the purposes of this lab you will need to install (if haven't already) and load the following packages.

```{r packages_lab4, eval = FALSE, echo = TRUE}
install.packages(c("vegan","ape","ggdendro","seqinr"))

#lapply is part of the apply family of functions, one of those functions that act as a for loop
lapply(c("vegan","ape","ggdendro","seqinr",
         "Biostrings", "tidyverse"),library, character.only = TRUE)

```

```{r packages2_lab4, echo = FALSE, warning=FALSE, message=FALSE,results='hide', include=FALSE}

#lapply is part of the apply family of functions, one of those functions that act as a for loop
lapply(c("vegan","ape","ggdendro","seqinr",
         "Biostrings", "tidyverse","ggrepel","DT"),library, character.only = TRUE)


```


## **Distances**{-}

When handling biological data distance measures, also known as dissimilarity measures or dissimilarity indices, 
are useful in comparing multiple entities. These entities can be populations (think allele dynamics), environmental
samples, gene sequences, genomes, gene expression data etc.
In essence every time we draw a comparison, we ask the quantitative question how 
much is X similar to Y or better yet how much are X and Y dissimilar.
The degree of dissimilarity between X and Y can simply be represented as percent dissimilarity or
might require a more sophisticated mathematical transformation. The approach used to measure 
dissimilarities depends on the nature of the data, it's format and of course the question in need of answering.
For example when performing a pairwise sequence alignment the alignment score can be thought of as
a relative measure of the dissimilarity between the two sequences and can be expressed as percent of the maximal alignment score.

When calculating distances between entities (e.g. populations) the data you work with can be tabulated, 
e.g. allele/genotype frequency/abundance table etc or non tabulated e.g. sequence alignment.
R can do both and the approaches are quite similar.  

### **Distance calculations with tabulated data**{-}

Base R is equipped to handle distance calculation with tabulated data. First let's generate data to
work with, by uploading the dune data-set. Dune is a part of the vegan package and it contains the counts of 30 different plants
(columns) from 20 different sites (rows). This would be similar to an allele/genotype/ gene
expression frequency data-set, only much smaller in scale, making it great training tool for the purpose of
this lab.

```{r load dune}
#Let's load the dune data set 
data("dune")
```

As you already know, there are multiple forms to store tabulated data in R.
Let's use the `class()` and `head()` function to figure out how dune is stored.

```{r check_dune, eval=FALSE}
#Now, lets figure out what we're dealing with
class(dune)
#Let's see how the table is structured
head(dune)
```

```{r check_dune2, echo=FALSE}
#Now, lets figure out what we're dealing with
class(dune)
#Let's see how the table is structured
DT::datatable(head(dune))
```

Note that the row names contains only numbers, it will make things clearer in the following steps if 
we modify samples names to be more meaningful. The code snippet bellow will use
the function `rownames()` to add S to each row which will denote site/sample.
In the `paste()` function I am asking R to generate a vector containing a sequence of S1-S20.
Alternatively, we could have used the tidyverse approach and added a column with sample identifier.
However, since we will be doing some calculations on the data in the data frame, it's easier to store the descriptive info in the row name than remembering to 
remove said column prior to the distance calculations.  

```{r modefy dune, eval=FALSE}
rownames(dune) <- paste("S",1:20,sep = "")
dune
```

```{r dune rn, echo=FALSE}
rownames(dune) <- paste("S",1:20,sep = "")
DT::datatable(dune)
```

***
>Note that I have used some hard coding in the snippet above.
>Can you think of waysto make it less hard coded?

***

#### **Using symmetric distances**{-}

Next let's calculate the distance between each of the sites using the `dist()` function from the stats
package, which is part of base R. When we calculate distances of tabulated data we are conducting a
pairwise comparison of every row on the table to every other row, based on the values found in each
of the columns. Rows with similar values in each column will receive a low score. The score represents distance, low 
score = short distance meaning that they are not very dissimilar (or very similar). In the snippet bellow I am using 
[euclidean distance measure](https://en.wikipedia.org/wiki/Euclidean_distance) to calculate the distance between each 
of the samples. Euclidean distances are considered symmetrical, I will explain why in the next segment.

```{r dist}
#Let's use the dist function to calculate the euclidean dist
dune.dist.euc <- dist(dune, method = "euclidean")

#The class of the variable created by the dist function is dist
class(dune.dist.euc)

#Lets print is
dune.dist.euc
```

The output looks like a distance table from an old road atlas, and that's exactly how it is read.
The distance between S1 and S2 is ~10.58, the distance between S1 and S12 is ~15.20, S1 is closer (or more similar, or less dissimilar)
to S2 than to S12 with respect to what species of plants are present in these site.

dist objects look like matrices but they are not, they are in fact oddly shaped numeric vectors.

```{r dist2}
#Dist objects are vectors
is.atomic(dune.dist.euc)

typeof(dune.dist.euc)
```



#### **Quick R exercise - working with dist objects**{-}

1. Use the `str()` function to study `dune.dist.euc`

2. What attributes does it have? Which aspect of the output from the `str()` function helps in 
reproducibility?

3. Print elements number 1, 2, and 20 from `dune.dist.euc`, what does that tell you about the order 
elements are indexed in dist objects?


#### **Using asymmetrical distances**{-}

Distance measures included in the dist functions are not necessarily designed to be used with 
biological data, since most of them are symmetric. Symmetric dissimilarity measures regard double
absence similarly to double presence. 

Say two populations show presence of a certain genotype or allele, dissimilarity measures will 
count this double presence towards lowering the dissimilarity score. Now say these populations also
show an absence of a certain allele or genotype, symmetric measures will treat this as a similarity
between the two populations as well. However, when it comes to biological data, the absence can
come from insufficient sampling of the population rather than true absence. Therefore need to use
an asymmetric measure. A measure that does not treat double absence and double presence equally. 

The R package vegan contains additional asymmetric dissimilarity measures such as Bray-Curtis.
One of the contributes to the package is [Dr. Gavin Simpson](https://fromthebottomoftheheap.net/)
a prof from the University of Regina, worth checking out if you are interested in ecological and environmental
informatics 


The snippet bellow uses the function `vegdist()`, the vegan implementation of the dist function, with 
[Bray-Curtis](https://en.wikipedia.org/wiki/Bray%E2%80%93Curtis_dissimilarity) as a dissimilarity index. 
The function is run twice, with binary set to either T or F.
When binary is set to T all values larger than 0 are converted to 1 and the rows are evaluated
based on presence and absence alone. When binary is set to F the distance factors in weight as well (i.e. not
just present but how much is present). These sort of measures are important in
molecular evolution but also when analyzing omics data (and ecology of course). 

```{r vegdists}
dune.vd.bry <- vegdist(dune, method = "bray", binary = F)

dune.vd.bry.bin <- vegdist(dune, method = "bray", binary = T)
```

#### **Quick R exercise - working with `vegdist()`**{-}

1. What class of objects does `vegdist()` output?

2. Use the `str()` function to explore `dune.vd.bry`

3. What will setting the arguments upper and diag to TRUE produce?

4. What other distance indices does `vegdist()` accepts under the method argument?

### **Two dimensional representation of distances**{-}

Distance matrices (or vectors) are multidimensional, and as such it becomes harder to draw
meaningful conclusions from them as the number of rows in your original table increase.

Ordination technique such as NMDS are helpful in dimension reduction and can be used to represent
the matrix in two dimensional space. We will not go into the details of ordination techniques in 
this course, just use them. However, in general all of them try to represent the distance matrix
accurately in as fewer dimensions as possible.

The snippet bellow will use vegan's metaMDS with the default setting of 2 dimensions. The end 
result of it will be each sample (line in the original dataframe) receiving two coordinates 
representing its location in two dimensional space and how it relates to every other sample.

```{r metaMDS1, results='hide'}

#Let's perform NMDS on the Bray-Curtis and the euclidean distances
#NMDS is a randomized iterative proceed, so setting seed to make sure the figure at the end looks
#the same for all of us.

#nmds for the euclidean distance matrix
set.seed(1)
dune.euc.nmds <- metaMDS(dune.dist.euc)

#nmds for the Bray-Curtis distance matrix
set.seed(1)
dune.bry.nmds <- metaMDS(dune.vd.bry)


```

The function `metaMDS()` can also call the `vegdist()` function and perform the distance calculation.


```{r metaMDS2, results='hide'}

#nmds with euc
set.seed(1)
dune.euc.nmds2 <- metaMDS(dune, distance = "euclidean")

#nmds with bray, note binary can be set here as well
set.seed(1)
dune.bry.nmds2 <- metaMDS(dune, distance = "bray", binary=F)

```



***

**quick R exercise**

1. What class of objects does dune.bry.nmds belong to?

2. What type of an object is it?

3. Explore the structure of dune.bry.nmds (hint, which function does that for you?)

4. How would you access the 3rd element of dune.bry.nmds?

5. What type of an object is the 3rd element of dune.bry.nmds?


***

If you completed the above exercise you saw that the output of metaMDS is essentially a list with
35 elements. What each element represents is not that important for this lab but in general the
list stores all the information needed to ensure the reproducibility of the output as well as
quality measures letting you know how good your results are, from a statistical standpoint.

For the sake of this lab we are interested in the coordinates of each sample and these are stored 
in the element points. Points is a matrix with two column of coordinates. Let's extract it and turn
it into a dataframe.

```{r extract points}

#extracting the points matrix as dataframe
#turning the row names into a column, note that I am doing it the spartan way and not using the 
#rownames_to_columns function from the dplyr package
#You can choose which ever method you preffer 


dune.euc.points <- as.data.frame(dune.euc.nmds$points) %>% 
  mutate(Samp = base::rownames(as.data.frame(dune.euc.nmds$points))) %>%
  select(Samp, everything())

dune.bry.points <- as.data.frame(dune.bry.nmds$points) %>% 
  mutate(Samp = base::rownames(as.data.frame(dune.bry.nmds$points))) %>%
  select(Samp, everything())

```


***

**Quick R exercise**

1. Use the head function to inspect each of the data frames

2. Do you find it useful in deciphering the relationships between the samples? 

***

The coordinates table might be simpler to look at compared to the distance matrix but it still
doesn't really help us understand which samples are more similar to one another.
In order to do that we must plot the samples using a scatter plot. 
Samples that appear close to each other in the plot, are more similar while distant samples are less similar.

Base R's plot function is a quick and dirty approach to generate scatter plot straight from the output
list.

```{r plot nmds base}
plot(dune.euc.nmds$points)

```

As always, base graphic is very spartan and not very informative. Let's use ggplot2 to generate a more informative
plot.


```{r nmds ggplot, fig.show='hide', collapse=TRUE}

#geom_point produces a scatter plot
p.euc <- ggplot(dune.euc.points, aes(x=MDS1, y = MDS2))+
  geom_point(size=2)+
  theme_bw(base_size = 16)+
  ggtitle("NMDS Dune dataset\neuclidean distances")

p.euc


p.bry <- ggplot(dune.bry.points, aes(x=MDS1, y = MDS2))+
  geom_point(size=2)+
  theme_bw(base_size = 16)+
  ggtitle("NMDS Dune dataset\nBray-Curtis distances")

p.bry




```

```{r prit p, echo=FALSE}

gridExtra::grid.arrange(p.euc,p.bry, ncol = 2)

```

Examining the two figures we can see that the two distance measures produced different 
configuration of the samples in two dimensional space. This was expected since we are comparing two
different approaches to measuring similarities between samples.
However, this does not really tell us how the configuration is different. One of the geometries of ggplot 2 is geom_text that allows for printing labels. However it requires adjustments of the coordinates to make sure that the points and labels do not overlap, which can be annoying. Luckily the ggrepel package offers additional geometries that does the dirty work for you. 

```{r ggrepel, echo=TRUE, eval=FALSE}
install.packages("ggrepel")

library(ggrepel)

```

Now, let's use geom_text_repel to label each point with it's sample name and compare the output of 
the two measures.

```{r ggplot2, fig.show='hide', collapse=TRUE}

p.euc2 <- ggplot(dune.euc.points, aes(x=MDS1, y = MDS2))+
  geom_point(size=2)+
  geom_text_repel(aes(label=Samp))+
  theme_bw(base_size = 16)+
  ggtitle("NMDS Dune dataset\neuclidean distances")

p.euc2

p.bry2 <- ggplot(dune.bry.points, aes(x=MDS1, y = MDS2))+
  geom_point(size=2)+
  geom_text_repel(aes(label=Samp))+
  theme_bw(base_size = 16)+
  ggtitle("NMDS Dune dataset\nBray-Curtis distances")

p.bry2





```

```{r p print2, echo=FALSE}

gridExtra::grid.arrange(p.euc2,p.bry2, ncol = 2)

```

Now we have a plot we can use to compare the two distance indices. We can see that not only does each method result in
a different configuration but also changes the way samples relate to each other. 
For example, using euclidean distances sample S1 appears to be more similar to sample 7 than sample S2 is to sample 7.
However, this is not the case when using Bray-Curtis. 

Dimension reduction and ordination techniques are great at visualizing the clustering of items into
groups, especially if large number of samples are involved. This is very useful when comparing a
large number of populations with respect to allele or genotype dynamics or when studying transcription or expression
patterns of a large number of individuals. 
<br>

***

>  **A note about distances and ordination techniques**
> 
As I mentioned, distance calculation ordination techniques (and hierarchical clustering which we
will deal with later in this lab) are important methods in exploring, understanding, and visualizing
data. You will use these a lot as bioinformaticians and it is imperative that you choose the
correct method for the data you are studying. A great resource for that is the book 
[Numerical ecology with R](http://adn.biol.umontreal.ca/~numericalecology/numecolR/). While the book deals mainly with ecological data, it is relevant to biology in general. Unfortunately, too often researches use approaches that
are unsuitable for the way their data is collected or distributed. This book will help you avoid such mistakes.
(Also, it's always great to plug the work of fellow Canadians!)

***

<br>

### **Distances with non-tabulated data**{-}

Non-tabulated data referrers mainly to sequence alignments, we will deal with multiple sequence 
alignment in lab 6. For this lab we will focus on understanding what object type results from 
comparing two or more sequences.
In most cases you will generate this sort of data in R (we will do that in lab #6), however 
sometimes you will have to import such data. Both Biostrings and seqinr have 
functions allowing them to read or accept alignment files or objects.

Let's first use Biostrings since the XString family of classes is very convenient to work with.

```{r biostrings msa}
#The file practice_al.fa contains an alignment of amino acis sequences
#let's read it using the dedicated function from Biostrings

al_bios <- readAAMultipleAlignment(
  filepath = "https://raw.githubusercontent.com/idohatam/Biol-3315-files/main/practice_al.fa",
  format = "FASTA")

```

***

**Quick R exercise**

1. What class is al_bios?

2. How many sequences does practice_al.fa contain?

3. Subset the first 10 sequences in the alignment

***

Now let's do the same with the dedicated function from seqinr. Note that seqinr is not fussy and 
will read any alignment with just one function, while Biostrings has a function for each type of
sequence. Did I mention seqinr is spartan?

```{r seqinr msa}
al_snr <- read.alignment(file = "practice_al.fa", format = "FASTA")

```

***

**Quick R exercise**

1. What class of an object is al_snr?

2. What type of an object is al_snr?

***

While seqinr's objects are less fancy than those of Biostrings, the seqinr package does have some
nice perks. One of them is calculating the distances between the sequences compared using the
sequence alignment.

```{r seqinr dist}

d_al_snr <- dist.alignment(al_snr)

```

***

**Quick R exercise**

1. What class is d_al_snr?

2. What is the structure of d_al_snr?

3. Print d_al_snr, does it remind you of any other object you worked with during the current lab?

***


While it is possible to reflect the distance matrix resulting from the alignment using 
ordination techniques it is more useful to visualize them using hierarchical visualization
tools such as dedrograms and phylogenetic trees, which is what we will do next. 


## **Dendrograms**{-}

BaseR has two classes of objects to house hierarchical information hclust and dendrogram.
The end result of both is a dendrogram, a tree like graph where similar objects (samples, sequences,
etc...) group together into clusters. Like 2D scatter plots, they are useful in fields outside 
evolution and phylogenetics. 

The hclust function performs hierarchical clustering of distance on the distance matrices you 
generated earlier using an assortment of aglomerative clustering algorithms.
We will not go into details of each clustering algorithm at this point, but will use this to study
the structure of the hclast object.

```{r hclust}

#Let's cluster the sequences from the multiple sequence alignment and see how do they relate to 
#each other

hc <- hclust(d_al_snr, method = "average")

#calling an hclust variable prints information used to generate the object
hc

#The output is of class hclust
class(hc)

#hclust variable is a type of a list (as always)
typeof(hc)


```

***

**Quick R exercise**

1. Use the str function to study hc

2. Convert the distance matrices generated for the dune data-set to an hclust object

***

As you saw hc contains 7 elements. Merge, hight and order are used to specify the topology of the
dendrogram, labels specify the label of each leaf, and the rest store information that helps with
documentation and reproducibility.

Class hclast can be converted to a class dendrogram using the as.dendrogram function.

```{r convert to dendro}
#Let's convert 
hcd <- as.dendrogram(hc)

```

***

**Quick R exercise**

1. What type of an object is the variable hcd

2. Call the variable hcd

3. Use the str function to study the structure of the variable hcd

4. Based on the variable call and the output of str how are variables of class hclust differ from
class dendrogram? Which class stores more information? 

5. Convert the hclust objects generated for the dune data-set to class dendrogram

***

### **Plotting dendrograms**{-}

Both class hclust and class dendro can be plotted using baseR plot function. 
This time the result is not that shabby.

```{r plot hc, collapse=TRUE, fig.show='hold', out.width='50%'}

#Let's plot the objects generated for the multiple sequence alignment
#xlab and ylab are axis titiles, main is the figure title. 
#Feel free to play with them


plot(hc, xlab="",ylab="", main = "hclust")

#par sets the parameters of the plot, mar sets the margins in the following order:
#bottom, left, top, right.
#Here I changed the setting of the bottom margin from it's default 5 to 9.7 to make sure that the 
#labels of the dendrogram are not cut.
par(mar=c(9.7,4,4,1)+0.1)

plot(hcd, xlab="",ylab="", main = "dendrogram")

```

As mentioned, hierarchical clustering groups similar elements together. In the case of the dendrogram
we plotted, sequences. Each vertical line in the dendrogram is referred to as a branch, each tip is
referred to as a leaf (in this case, each sequence is a leaf, when you plot the dune data-set each
sample will be a leaf). While the length of the branches between the two plots might be different
they are read the same, sequences that branch from the same junction are more similar to each other
than the rest. For example the sequence A9MLE0 and !83J28 are more similar to each other than to
any other sequence. The level of dissimilarity between them is denoted by the scale of the y axis
(0.2) which is the scale of the distance measure used in the clustering process, here Hamming
distance. Similarly both these sequences are equally distant from A7FP80, and both are the closest
sequences to it.

***

**Quick R exercise**

1. Which sequences are the closest to the sequence A0A378D0M5?

2. What is the level of dissimilarity between the group of 4 sequences to the left and the rest of
the sequences?

***

From the figures we can see that the sequences cluster into at least two groups (left and right).
In lab 6 (maybe) and in advanced courses (e.g. all the omics courses) we will use some statistical
approaches to test what is the optimal number of groups that the leafs cluster too.



***

**Quick R exercise**

1. Plot the class dendrogram (no need to plot class hclast) variables you prepared for the dune
data-set

2. Do you notice any difference in the topology of the trees, order of branches between the one based on euclidean distances and the one based on non-binary Bray-Curtis distances?

***

In lab 6 and 7 we will use the package dendextend to directly compare the topology of the different
trees and dendrograms we generate.


#### **Plotting dendrograms with ggplot2**{-}

ggdendro is a package the converts dendrogram class variable to a format that can be used by 
ggplot2 to plot dendrograms. This can be very useful due to the flexibility ggplot2 offers in plot
design and output. Furthermore, the resulting class stores data in a way that is easily accessed
and modified using other tidyverse packages such as dplyr and tidyr.


Variables of class dendrogram can be converted to class dendro, the class type used by ggdendro,
using the dendro_data function.

```{r ggd1}

#let's convert hcd to class dendro
hcggd <- dendro_data(hcd)

class(hcggd)


```


***

**Quick R exercise**

1. Use the str function to study hcggd

2. What type of an object is hcggd?

3. What type of an object are the elements labels and segments?

4. What type of information do you think is stored segments?

***

ggdendro class variables, as you noticed, contain two important dataframes, segments and labels.
The segment dataframe contains the topology of the tree, and consists of 4 columns.


```{r segments, echo=T}

head(hcggd$segments)

```


x and xend denote the horizontal position of each branch of the tree (i.e. its coordinates along
the x axis). y and yend denote the vertical positions of each branch of the tree (i.e. its
coordinates along the y axis).

The label dataframe contains the labels of each of the leaves (shocking I know), and consists of 3 columns.   

```{r labels, echo=T}

head(hcggd$labels)

```


x denote the position along the x axis and y denotes the position along the y axis which is always
0. label contains...you got it. Now that we understand the structure of the dendro class we can  plot dendrograms with it using the ggplot2 package.


```{r ggd2}

#Let's plot the dendrogram with ggplot2
#The dendrogram has two parts, the segments which are basically lines starched between two points 
#in two dimensional space, and the labels. 
#That's why I did not specify the x and y coordinates in the initial ggplot call, but rather 
#specified it for each of the geoms
p <- ggplot(data = hcggd$segments) + 
  #geom_segment draws the segments (branches/lines) of the dendrogram
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) + 
  #geom_text adds a text layer to the plot, which are the leaves labels
  #the angle argument sets the orientation of the text and the hjust makes sure that the 
  #labels are justified horizontal
  #within the aes y is set to y-0.01 so that the labels will be a bit further away from the tips of
  #each leaf
  geom_text(data = hcggd$labels, aes(x, y-0.01, label = label),
            hjust = 1, angle = 90, size = 4)+
  theme_bw(base_size = 16)+
  #Don't really need any ticks or text for the x axis
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())+
  #the ylim function adjusts the range of values that the ggplot includes in the plot
  #here I set it to -0.45 - 0.62 to make sure none of the segments or labels get cut off
  #The default would cut it after -0.2 which means that some of the labels would not be 
  #fully displayed
  ylim(min(hcggd$segments$y)-1, max(hcggd$segments$yend))+
  xlab(NULL)+
  ylab(NULL)
p


```

I sold you on the fact that ggplot makes things flexible and great but so far it looks more or less
like baseR graphic. Lets see an example of that. 
As I mentioned, there are several mathematical methods (e.g. k-means clustering) that help calculate the optimal
number of clusters produced by the hierarchical clustering. 

For the following graph assume a hypothetical scenario where k-means indicates 
that the optimal way to group the sequences would be into 3 groups, the four on the left would be group 1, the four 
sequences of the right would be group 3 and the rest are group 2. 
It is easy to add a fourth column to each of the dataframes of the class dendro object with that
information and then overlay the grouping on the dendrogram.


```{r ggd3}

#Use mutate to add a column named Group which contains which group each sequence belongs to
#We can use the positions of the labels to parse that, labels with position along the x axis smaller than 5 belong
#the 4 sequences on the left so we mark them as group 1, will do the same for the other two groups. 
#Can also use case_when.

hcggd$labels <- hcggd$labels %>%
  mutate(Group = if_else(x < 5, "1", "2")) %>%
  mutate(Group = if_else(x > 11, "3", Group))


#Use a similar process with the segments
hcggd$segments <- hcggd$segments %>%  
  mutate(Group = if_else(x < 5, "1", "2")) %>%
  mutate(Group = if_else(x > 11, "3", Group))

#Now let's plot the dendrogram
p2 <- ggplot(data = hcggd$segments) +
  #Note that I color by the Group column
  #alpha denotes how transparent the coloring will be, the default is 1 and I want the colors to be
  #more gentle
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend, color=Group), 
               size = 1.25, alpha=0.5) + 
  geom_text(data = hcggd$labels, aes(x, y-0.01, label = label, color = Group),
            show.legend = F,
            hjust = 1, angle = 90, size = 4, alpha=0.5)+
  theme_bw(base_size = 16)+
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = c(0.93,0.89),
        legend.background = element_blank())+
  #Setting the colors manually because the red and blue default are ugly
  scale_color_manual(values = c("dark red", "black", "dark green"))+
  ylim(min(hcggd$segments$y)-1, max(hcggd$segments$yend))+
  xlab(NULL)+
  ylab(NULL)
p2


```

<br>

***

> **A note about machine learning**
>
While this plot might not look impressive, what you have done here is to take your first steps
(unless you've done this before) in machine learning. When you ask an algorithm (e.g. k-means
clustering) to classify your samples into groups, without specifying groups, you are performing an
unsupervised machine learning. 
Once you have enough information about how the particular set of data classifies, you can use it to
develop a predictive algorithm that would classify new samples into groups based on the information
resulting from the unsupervised learning. When you train or use an algorithm to predict which
category a sample would be classified to you are performing supervised machine learning.

***

<br>


## **Phylogenetic trees**{}

Phylogenetic trees are tree like graphs that not only denote relationship between each leaf on the
x axis, but also include a time component on the y axis (i.e. junctions lower on the tree are
assumed to have occurred earlier than those showing upper on the tree). Phylogenetic trees require 
to store more information than dendrogram (e.g. information about taxa, molecular information, 
genetic codes, model assumptions etc). The package ape offers the class phylo which stores 
phylogenetic trees.

There are many methods to generate pylogenetic trees and several file formats in which
phylogenetic trees are written ( [nexus](https://en.wikipedia.org/wiki/Nexus_file),
[newick](https://en.wikipedia.org/wiki/Newick_format) are the most common file formats ), we will
go over those during lab 7. However at the end of the day they all result in a phylo or phylo like classes of
variables. Here we will only get familiarized with said class. 

Because all tree like objects in R are stored as lists, it is easy to convert between
them. The snippet bellow uses the as.phylo command to generate a phylo object 

```{r phylo1}

hcp <- ape::as.phylo(hc)
hcp

```

***

**Quick R exercise**

1. Explore the list outputted by as.phylo (hcp) which elements make up this list?

2. What class does each element belong to?

***

As you found out, hcp contains 4 elements. The matrix edge consists of two columns

```{r phylo2, echo=FALSE}

head(hcp$edge)

```

Each row in the matrix represent a branch or an edge, if this sounds like terminology from graph
theory you are correct. Indeed ape treats trees almost like association networks and actually
allows you to add columns to the edge matrix and plot a network.

The numeric vector edge length contains the length of each branch of the tree

```{r phylo3, echo=FALSE}
head(hcp$edge.length)

```

It's order corresponds to the order of rows in the edge matrix.

The character vector tip.label surprisingly contains the label of each leaf of the tree, making
those easy to modify (for example remove the UniRef90_ tag).

Nnode represents the number of nodes in the tree again this is terminology borrowed from graph 
theory and represents each branching point along the tree (in our case 14).

As I mentioned phylo class objects can contain additional information about the tree. Additional
elements include: 

* node.label a character vector that contains node labels for example bootstrap values.

* root.edge a numeric value giving the length of the branch at the root if it exists

* tip.data a vector dating the tips (leaves) of the tree, useful if some are not current




### **Plotting phylogenetic trees**{-}

Plotting phylogenetic trees is straight forward using baseR. Examples include a node label element which contains labels of each branching point. 

```{r basep}

plot(hcp)


```


Alternatively trees can be unrooted.

```{r basep2}
plot(hcp, type = "unrooted")

```



Plotting phylogentic trees using ggplot2 requires converting the phylo object to dendro and 
following the workflow we used earlier.


## **Summary assignment**{-}

Start a script file in RStudio name it 3315_lab4_yourinitials.R
You will submit this script and output file if applicable to Brightspace by Tue October 18 23:59.

**1.** Convert the distance matrices generated for the dune data-set (euclidean, and both Bray-Curtis) to an hclust object.

**2.** Convert the distance matrices generated for the dune data-set to an R matrix (you should have 3 matrices) and save them as CSV files. Make sure the file names are descriptive.

**3.** Convert the class hclast variables you have created to class dendrogram

**4.** Plot both the class dendrogram and class hclass variables you created for the dune data-set using base R, give title using the argument main. Save plots as jpeg using the RStudio plots tab.

  + Do you see any difference in clustering between each of the distance measures?

**5.** Convert the dendrogram objects to dendro objects and plot using ggplot 2

**6.** Export plots as pdf using ggsave

**7.** Add the column altitude to the label dataframe that's part of the dednro class variables, have the values "low", "high", "sea level" added to each sample randomly.

**8.** Add the column Groups to the segment dataframe of the dendro class variables you generated. 
Divide the two Bray-Curtis based dendrograms to 3 groups and the euclidean to 2 groups based on the
order of the samples.

**9.** Plot each of the dendro class objects, color the labels based on altitude and the branches based on the groups column. If this was real data would you see any trends in clustering of these samples based on altitude? Make sure to export these plots as pdfs as well

**10.** Use the ?write.tree and ?write.nexus to export the phylogenetic tree generated in this lab. Give a descriptive file name so that it is clear which is a nexus and which is a newick format.

  + Which functions would you have used to import each of the files you exported?
  
**11.** Write a function that reads all csv files from a folder, log transforms them and uses metaMDS to perform NMDS on all files with binary bray as distance. Make sure that the results are all stored as element of a list (hint make sure that the list has the same number of elements as the number of files in the folder). Make sure that your function returns said list.

