```{r setup}
#| include: false
knitr::opts_chunk$set(echo = TRUE, comment = "", prompt = TRUE, class.source = "fold-show")
options(knitr.duplicate.label = "allow")
```

```{r klippy}
#| echo: false
#| include: true
klippy::klippy(position = "right")
```

<br>

# **Lab 4: Handling of Biological Data in R - 2**

<br>

## **Learning objectives** 
<details>
  <summary>Click to expand</summary>
__By the end of this lab you will be able to:__


*	Demonstrate the ability to generate distance matrices from a variety of data types

* Plot distances 

* Demonstrate the ability to generate dendrograms 

* Demonstrate the ability to generate phylogenetic trees

* Demonstrate the ability to export and import trees and dendrograms

* Demonstrate the ability to plot trees and dendrograms

</details>

***

## **Lab 4**

In the previous lab you were introduced to handling biological strings with R. In this lab you 
will learn how to handle some of the output formats that result from analyzing those strings in R.
In most cases, as you will see in lecture and subsequent labs, this entails working with distance 
matrices and some sort of hierarchical clustering of those matrices.

R's basic data structures, mainly lists, provide very flexible containers to house such complex types
of information. As you will see, most non-bioconductor packages house their outputs in list form or
at least as list-based objects.

For the purposes of this lab you will need to install and load the following packages.

```{r packages}
#| eval: false
#| echo: true
install.packages(c("vegan","ape","ggdendro","seqinr"))

# lapply is part of the apply family of functions, one of those functions that act as a for loop
lapply(c("vegan","ape","ggdendro","seqinr",
         "Biostrings", "tidyverse"),library, character.only = TRUE)
```

```{r packages2}
#| echo: false
#| warning: false
#| message: false
#| results: hide
#| include: false
lapply(c("vegan","ape","ggdendro","seqinr",
         "Biostrings", "tidyverse","ggrepel"),library, character.only = TRUE)
```


## **Distances**

When handling biological data, distance measures—also known as dissimilarity measures or indices—
are useful in comparing multiple entities. These entities can be populations, environmental
samples, gene sequences, genomes, expression data, etc.
In essence, every time we draw a comparison, we ask the quantitative question: how 
similar is X to Y, or better yet, how dissimilar are X and Y?
The degree of dissimilarity between X and Y can simply be represented as percent dissimilarity or
might require more sophisticated mathematical transformation. The approach used to measure 
dissimilarities depends on the nature of the data, its format, and of course the question in need of answering.
For example, when performing a pairwise sequence alignment, the alignment score can be thought of as
a relative measure of the dissimilarity between the two sequences and can be expressed as a percentage of the maximal alignment score.

When calculating distances between entities (e.g., populations), the data you work with can be tabulated (e.g., allele/genotype frequency/abundance table, etc.) or non-tabulated (e.g., sequence alignment).
R can do both, and the approaches are quite similar.  

### **Distance calculations with tabulated data**

`BaseR` is equipped to handle distance calculation with tabulated data. First let's generate data to
work with by uploading the dune dataset that's part of the `vegan` package.
This dataset contains the counts of 30 different plants
(columns) in 20 different sites (rows). This would be similar to an allele/genotype/gene
expression frequency dataset, only much smaller, making it a great training tool for the purpose of
this lab.

```{r load dune}

# Let's load the dune dataset 

data("dune")

# Now, let's figure out what we're dealing with
class(dune)

# Let's see how the table is structured
head(dune)

```

Note that the row names are made of numbers. It will make things clearer in the following steps if 
we modify sample names to be more meaningful. 

```{r modify dune}
# Let's add S to each row, denoting site or sample.
# Reminder, R is case sensitive
# Note that here I am asking paste to generate a vector with a sequence of numbers 1-20
# Then paste S to each element in the vector with no space separation
# Then assign the new character vector S1-S20 to the rownames of the dataframe
rownames(dune) <- paste("S",1:20,sep = "")

```

***

**Quick R exercise**

The previous code snippet contained some hard coding. Can you think of two ways
to make it less hard coded?

***

Next let's calculate the distance between each of the sites using the `dist()` function from the stats
package, which is part of `baseR`. When we calculate distances of tabulated data, we are conducting a
pairwise comparison of every row in the table to every other row, based on the values found in each
of the columns. Rows with similar values in each column will receive a low score. The score represents distance; low 
score = short distance, meaning that they are not very dissimilar (or very similar). In the snippet below I am using 
[euclidean distance measure](https://en.wikipedia.org/wiki/Euclidean_distance) to calculate the distance between each 
of the samples.

```{r dist}

# Let's use the dist function to calculate the euclidean distance
dune.dist.euc <- dist(dune, method = "euclidean")

# The class of the variable created by the dist function is dist
class(dune.dist.euc)

# Let's print it

dune.dist.euc



```

The output looks like a distance table from an old road atlas, and that's exactly how it is read.
The distance between S1 and S2 is ~10.58, the distance between S1 and S12 is ~15.20. S1 is closer
to S2 than S12 with respect to what species of plants are present in this site and their
abundance.

Dist objects look like matrices but they are not; they are in fact oddly shaped numeric vectors.

```{r dist2}
# Dist objects are vectors
is.atomic(dune.dist.euc)

typeof(dune.dist.euc)

```


***

**Quick R exercise**

1. Use the `str()` function to study dune.dist.euc

2. What attributes does it have? Which aspect of the output from the `str()` function helps with 
reproducibility?

3. Print elements number 1, 2, and 20 from `dune.dist.euc`. What does that tell you about the order 
elements are indexed in dist objects?

***

Distance measures included in the `dist()` function are not necessarily designed to be used with 
biological data, since most of them are symmetric. Symmetric dissimilarity measures regard double
absence similarly to double presence. 

Say two populations show presence of a certain genotype or allele; dissimilarity measures will 
count this double presence toward lowering the dissimilarity score. Now say these populations also
show an absence of a certain allele or genotype. Symmetric measures will treat this as a similarity
between the two populations as well. However, when it comes to biological data, the absence can
come from insufficient sampling of the population rather than true absence. Therefore, we need to use
an asymmetric measure—a measure that does not treat double absence and double presence equally. 

The R package `vegan` contains additional asymmetric dissimilarity measures such as Bray-Curtis.
One of the contributors to the package is [Dr. Gavin Simpson](https://fromthebottomoftheheap.net/),
a professor from the University of Regina, worth checking out if you are interested in ecological
informatics.


The snippet below uses the function `vegdist()`, the vegan implementation of the `dist()` function, with 
[Bray-Curtis](https://en.wikipedia.org/wiki/Bray%E2%80%93Curtis_dissimilarity) as a dissimilarity index. 
The function is run twice, with binary set to either T or F.
When binary is set to T, all values larger than 0 are converted to 1 and the rows are evaluated
based on presence and absence alone. When binary is set to F, the distance factors in weight as well (i.e., not
just present but how much is present). These sorts of measures are important in
molecular evolution but also when analyzing omics data (and ecology of course). 

```{r vegdists}

# Let's run vegdist
dune.vd.bry <- vegdist(dune, method = "bray", binary = F)

dune.vd.bry.bin <- vegdist(dune, method = "bray", binary = T)

```

***

**Quick R exercise**

1. What class of objects does `vegdist()` output?

2. Use the `str()` function to explore `dune.vd.bry`

3. What will setting the arguments `upper` and `diag` to `TRUE` produce?

4. What other distance indices does `vegdist()` accept under the method argument?


***


### **Two dimensional representation of distances**

Distance matrices (or vectors) are multidimensional, and as such it becomes harder to draw
meaningful conclusions from them as the number of rows in your original table increases.

Ordination techniques such as NMDS are helpful in dimension reduction and can be used to represent
the matrix in two-dimensional space. We will not go into the details of ordination techniques in 
this course, just use them. However, in general all of them try to represent the distance matrix
accurately in as few dimensions as possible.

The snippet below will use vegan's `metaMDS()` with the default setting of 2 dimensions. The end 
result of it will be each sample (line in the original dataframe) receiving two coordinates 
representing its location in two-dimensional space and how it relates to every other sample.

```{r metaMDS1}
#| results: hide

# Let's perform NMDS on the Bray-Curtis and the euclidean distances
# NMDS is a randomized iterative process, so setting seed to make sure the figure at the end looks
# the same for all of us.

# nmds for the euclidean distance matrix
set.seed(1)
dune.euc.nmds <- metaMDS(dune.dist.euc)

# nmds for the Bray-Curtis distance matrix
set.seed(1)
dune.bry.nmds <- metaMDS(dune.vd.bry)


```

The function `metaMDS()` can also call the `vegdist()` function and perform the distance calculation.


```{r metaMDS2}
#| results: hide

# nmds with euclidean
set.seed(1)
dune.euc.nmds2 <- metaMDS(dune, distance = "euclidean")

# nmds with bray, note binary can be set here as well
set.seed(1)
dune.bry.nmds2 <- metaMDS(dune, distance = "bray", binary=F)

```



***

**Quick R exercise**

1. What class of objects does `dune.bry.nmds` belong to?

2. What type of an object is it?

3. Explore the structure of `dune.bry.nmds` (hint: which function does that for you?)

4. How would you access the 3rd element of `dune.bry.nmds`?

5. What type of an object is the 3rd element of `dune.bry.nmds`?


***

If you completed the above exercise, you saw that the output of `metaMDS()` is essentially a list with
35 elements. What each element represents is not that important for this lab, but in general the
list stores all the information needed to ensure the reproducibility of the output as well as
quality measures letting you know how good your results are from a statistical standpoint.

For the sake of this lab, we are interested in the coordinates of each sample, and these are stored 
in the element points. The points element is a matrix with two columns of coordinates. Let's extract it and turn
it into a dataframe.

Let's first extract the points for the euclidean distances:

```{r extract points euc}

# Extracting the points matrix as dataframe
# Turning the row names into a column; note that I am doing it the spartan way and not using the 
# rownames_to_columns function from the dplyr package
# You can choose whichever method you prefer 

dune.euc.points <- as.data.frame(dune.euc.nmds$points) %>% 
  mutate(Samp = base::rownames(as.data.frame(dune.euc.nmds$points))) %>%
  select(Samp, everything())

```

Next, we extract the points for the Bray-Curtis distances:

```{r extract points bry}

dune.bry.points <- as.data.frame(dune.bry.nmds$points) %>% 
  mutate(Samp = base::rownames(as.data.frame(dune.bry.nmds$points))) %>%
  select(Samp, everything())

```


***

**Quick R exercise**

1. Use the `head()` function to inspect each of the dataframes

2. Do you find it useful in deciphering the relationships between the samples? 

***

The coordinates table might be simpler to look at compared to the distance matrix, but it still
doesn't really help us understand which samples are more similar to one another.
In order to do that, we must plot the samples using a scatter plot. 
Samples that appear close to each other in the plot are more similar, while distant samples are less similar.

BaseR's `plot()` function is a quick and dirty approach to generate a scatter plot straight from the output
list.

```{r plot base}
plot(dune.euc.nmds$points)


```

As always, base graphics are very spartan and not very informative. Let's use `ggplot2` to generate a more informative
plot. First, we create a scatter plot for the euclidean distances:


```{r ggplot euc}
#| fig.show: hide
#| collapse: true

# geom_point produces a scatter plot
p.euc <- ggplot(dune.euc.points, aes(x=MDS1, y = MDS2))+
  geom_point(size=2)+
  theme_bw(base_size = 16)+
  ggtitle("NMDS Dune dataset\neuclidean distances")

p.euc

```

Next, we do the same for the Bray-Curtis distances:

```{r ggplot bry}
#| fig.show: hide
#| collapse: true

p.bry <- ggplot(dune.bry.points, aes(x=MDS1, y = MDS2))+
  geom_point(size=2)+
  theme_bw(base_size = 16)+
  ggtitle("NMDS Dune dataset\nBray-Curtis distances")

p.bry

```

Let's print the two plots together to compare them.

```{r prit p}
#| echo: false

gridExtra::grid.arrange(p.euc,p.bry, ncol = 2)

```

Examining the two figures, we can see that the two distance measures produced different 
configurations of the samples in two-dimensional space. This was expected since we are comparing two
different approaches to measuring similarities between samples.
However, this does not really tell us how the configuration is different. One of the geometries of `ggplot2` is `geom_text()` that allows for printing labels. However, it requires adjustments of the coordinates to make sure that the points and labels do not overlap, which can be annoying. Luckily, the `ggrepel` package offers additional geometries that do the dirty work for you. 

```{r ggrepel}
#| echo: true
#| eval: false
install.packages("ggrepel")

library(ggrepel)

```

Now, let's use `geom_text_repel()` to label each point with its sample name and compare the output of 
the two measures.

```{r ggplot2 euc}
#| fig.show: hide
#| collapse: true

p.euc2 <- ggplot(dune.euc.points, aes(x=MDS1, y = MDS2))+
  geom_point(size=2)+
  geom_text_repel(aes(label=Samp))+
  theme_bw(base_size = 16)+
  ggtitle("NMDS Dune dataset\neuclidean distances")

p.euc2

```

```{r ggplot2 bry}
#| fig.show: hide
#| collapse: true

p.bry2 <- ggplot(dune.bry.points, aes(x=MDS1, y = MDS2))+
  geom_point(size=2)+
  geom_text_repel(aes(label=Samp))+
  theme_bw(base_size = 16)+
  ggtitle("NMDS Dune dataset\nBray-Curtis distances")

p.bry2
```

```{r p print2}
#| echo: false
gridExtra::grid.arrange(p.euc2,p.bry2, ncol = 2)
```

Now we have a plot we can use to compare the two distance indices. We can see that not only does each method result in
a different configuration, but it also changes the way samples relate to each other. 
For example, using euclidean distances, sample S1 appears to be more similar to sample S7 than sample S2 is to sample S7.
However, this is not the case when using Bray-Curtis. 

Dimension reduction and ordination techniques are great at visualizing the clustering of items into
groups, especially if large numbers of samples are involved. This is very useful when comparing a
large number of populations with respect to allele or genotype dynamics or when studying transcription or expression
patterns of a large number of individuals. 
<br>

***

::: {.callout-note}
 **A note about distances and ordination techniques**

As I mentioned, distance calculation, ordination techniques, and hierarchical clustering (which we
will deal with later in this lab) are important methods in exploring, understanding, and visualizing
data. You will use these a lot as bioinformaticians, and it is imperative that you choose the
correct method for the data you are studying. A great resource for that is the book 
[Numerical ecology with R](http://adn.biol.umontreal.ca/~numericalecology/numecolR/). While the book deals mainly with ecological data, it is relevant to biology in general. Unfortunately, too often researchers use approaches that
are unsuitable for the way their data are collected or distributed. This book will help you avoid such mistakes.
(Also, it's always great to plug the work of fellow Canadians!)
:::


***

<br>

### **Non-tabulated data**

Non-tabulated data refers mainly to sequence alignments. We will deal with multiple sequence 
alignments in lab 6. For this lab, we will focus on understanding what object type results from 
comparing two or more sequences.
In most cases you will generate this sort of data in R (we will do that in lab 6), however, 
sometimes you will have to import such data. Both `Biostrings` and `seqinr` have 
functions allowing them to read or accept alignment files or objects.

Let's first use `Biostrings` since the `XString` family of classes is very convenient to work with.

```{r biostrings msa}
# The file practice_al.fa contains an alignment of amino acid sequences
# Let's read it using the dedicated function from Biostrings

al_bios <- readAAMultipleAlignment(
  filepath = "https://raw.githubusercontent.com/idohatam/Biol-3315-files/main/practice_al.fa", 
  format = "FASTA")

```

***

**Quick R exercise**

1. What class is `al_bios`?

2. How many sequences does `practice_al.fa` contain?

3. Subset the first 10 sequences in the alignment

***

Now let's do the same with the dedicated function from `seqinr`. Note that `seqinr` is not fussy and 
will read any alignment with just one function, while `Biostrings` has a function for each type of
sequence (AA, DNA, RNA). Did I mention `seqinr` is spartan?

```{r seqinr msa}
al_snr <- seqinr::read.alignment(
  file = "https://raw.githubusercontent.com/idohatam/Biol-3315-files/main/practice_al.fa", 
  format = "FASTA")

```

***

**Quick R exercise**

1. What class of an object is `al_snr`?

2. What type of an object is `al_snr`?

***

While `seqinr's` objects are less fancy than those of `Biostrings`, the `seqinr` package does have some
nice perks. One of them is calculating the distances between the sequences in the sequence alignment.

```{r seqinr dist}
d_al_snr <- dist.alignment(al_snr)
```

***

**Quick R exercise**

1. What class is `d_al_snr`?

2. What is the structure of `d_al_snr`?

3. Print `d_al_snr`. Does it remind you of any other object you worked with during the current lab?

***


While it is possible to reflect the distance matrix resulting from the alignment using 
ordination techniques, it is more useful to visualize them using hierarchical visualization
tools such as dendrograms and phylogenetic trees, which is what we will do next. 


## **Dendrograms**

`BaseR` has two classes of objects to house hierarchical information: `hclust` and `dendrogram`.
The end result of both is a dendrogram, a tree-like graph where similar objects (samples, sequences,
etc.) group together into clusters. Like 2D scatter plots, they are useful in fields outside 
evolution and phylogenetics. 

The `hclust()` function performs hierarchical clustering on the distance matrices you 
generated earlier using an assortment of agglomerative clustering algorithms.
We will not go into the details of each clustering algorithm at this point, but will use this to study
the structure of the `hclust` object.

```{r hclust}

# Let's cluster the sequences from the multiple sequence alignment and see how they relate to 
# each other

hc <- hclust(d_al_snr, method = "average")

# Calling an hclust variable prints information used to generate the object
hc

# The output is of class hclust
class(hc)

# hclust variable is a type of a list (as always)
typeof(hc)


```

***

**Quick R exercise**

1. Use the `str()` function to study `hc`

2. Convert the distance matrices generated for the dune dataset to an `hclust` object

***

As you saw, `hc` contains 7 elements. Merge, height, and order are used to specify the topology of the
dendrogram; labels specify the label of each leaf; and the rest store information that helps with
documentation and reproducibility.

Class hclust can be converted to a class dendrogram using the `as.dendrogram()` function.

```{r convert to dendro}
# Let's convert 
hcd <- as.dendrogram(hc)

```

***

**Quick R exercise**

1. What type of an object is the variable `hcd`?

2. Call the variable `hcd`

3. Use the `str()` function to study the structure of the variable `hcd`

4. Based on the variable call and the output of `str()`, how do variables of class hclust differ from
class dendrogram? Which class stores more information? 

5. Convert the hclust objects generated for the dune dataset to class dendrogram

***

### **Plotting dendrograms**

Both class hclust and class dendrogram can be plotted using the `plot()` function. 
This time the result is not too shabby. First, let's plot the `hclust` object.

```{r plot hclust}
#| collapse: true
#| fig.show: hold
#| out.width: 50%

# Let's plot the objects generated for the multiple sequence alignment
# xlab and ylab are axis titles, main is the figure title. 
# Feel free to play with them

plot(hc, xlab="",ylab="", main = "hclust")

```

Next, let's plot the `dendrogram` object.

```{r plot dendrogram}
#| collapse: true
#| fig.show: hold
#| out.width: 50%

# par sets the parameters of the plot; mar sets the margins in the following order:
# bottom, left, top, right.
# Here I changed the setting of the bottom margin from its default 5 to 9.7 to make sure that the 
# labels of the dendrogram are not cut.
par(mar=c(9.7,4,4,1)+0.1)

plot(hcd, xlab="",ylab="", main = "dendrogram")

```

As mentioned, hierarchical clustering groups similar elements together. In the case of the dendrogram
we plotted, sequences. Each vertical line in the dendrogram is referred to as a branch; each tip is
referred to as a leaf (in this case, each sequence is a leaf; when you plot the dune dataset, each
sample will be a leaf). While the length of the branches between the two plots might be different,
they are read the same: sequences that branch from the same junction are more similar to each other
than the rest. For example, the sequences A9MLE0 and I83J28 are more similar to each other than to
any other sequence. The level of dissimilarity between them is denoted by the scale of the y axis
(0.2), which is the scale of the distance measure used in the clustering process—here, Hamming
distance. Similarly, both these sequences are equally distant from A7FP80, and both are the closest
sequences to it.

***

**Quick R exercise**

1. Which sequences are the closest to the sequence A0A378D0M5?

2. What is the level of dissimilarity between the group of 4 sequences to the left and the rest of
the sequences?

***

From the figures, we can see that the sequences cluster into at least two groups (left and right).
In lab 6 and in advanced courses (e.g., all the omics courses), we will use some statistical
approaches to test what is the optimal number of groups that the leaves cluster into.



***

**Quick R exercise**

1. Plot the class dendrogram (no need to plot class hclust) variables you prepared for the dune
dataset

2. Do you notice any difference in the topology of the trees, order of branches between the one based on euclidean distances and the one based on non-binary Bray-Curtis distances?

***

In labs 6 and 7, we will use the package dendextend to directly compare the topology of the different
trees and dendrograms we generate.


**Plotting dendrograms with ggplot2**

`ggdendro` is a package that converts dendrogram class variables to a format that can be used by 
`ggplot2` to plot dendrograms. This can be very useful due to the flexibility `ggplot2` offers in plot
design and output. Furthermore, the resulting class stores data in a way that is easily accessed
and modified using other `tidyverse` packages such as `dplyr` and `tidyr`.


Variables of class dendrogram can be converted to class dendro, the class type used by ggdendro,
using the `dendro_data()` function.

```{r ggd1}

# Let's convert hcd to class dendro
hcggd <- dendro_data(hcd)

class(hcggd)


```


***

**Quick R exercise**

1. Use the `str()` function to study `hcggd`

2. What type of an object is `hcggd`?

3. What type of an object are the elements labels and segments?

4. What type of information do you think is stored in segments?

***

ggdendro class variables, as you noticed, contain two important dataframes: segments and labels.
The segment dataframe contains the topology of the tree and consists of 4 columns.


```{r segments}
#| echo: true

head(hcggd$segments)

```


x and xend denote the horizontal position of each branch of the tree (i.e., its coordinates along
the x axis). y and yend denote the vertical positions of each branch of the tree (i.e., its
coordinates along the y axis).

The label dataframe contains the labels of each of the leaves (shocking, I know) and consists of 3 columns.   

```{r labels}
#| echo: true

head(hcggd$labels)

```


x denotes the position along the x axis, and y denotes the position along the y axis, which is always
0. label contains...you got it. Now that we understand the structure of the dendro class, we can plot dendrograms with it using the `ggplot2` package.


```{r ggd2}

# Let's plot the dendrogram with ggplot2
# The dendrogram has two parts: the segments, which are basically lines stretched between two points 
# in two-dimensional space, and the labels. 
# That's why I did not specify the x and y coordinates in the initial ggplot call, but rather 
# specified them for each of the geoms
p <- ggplot(data = hcggd$segments) + 
  # geom_segment draws the segments (branches/lines) of the dendrogram
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) + 
  # geom_text adds a text layer to the plot, which are the leaf labels
  # The angle argument sets the orientation of the text and the hjust makes sure that the 
  # labels are justified horizontally
  # Within the aes, y is set to y-0.01 so that the labels will be a bit further away from the tips of
  # each leaf
  geom_text(data = hcggd$labels, aes(x, y-0.01, label = label),
            hjust = 1, angle = 90, size = 4)+
  theme_bw(base_size = 16)+
  # Don't really need any ticks or text for the x axis
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())+
  # The ylim function adjusts the range of values that ggplot includes in the plot
  # Here I set it to -0.45 - 0.62 to make sure none of the segments or labels get cut off
  # The default would cut it after -0.2, which means that some of the labels would not be 
  # fully displayed
  ylim(min(hcggd$segments$y)-1, max(hcggd$segments$yend))+
  xlab(NULL)+
  ylab(NULL)
p


```

I sold you on the fact that `ggplot2` makes things flexible and great, but so far it looks more or less
like `baseR` graphics. Let's see an example of that. 
As I mentioned, there are several mathematical methods (e.g., k-means clustering) that help calculate the optimal
number of clusters produced by the hierarchical clustering. 

For the following graph, assume a hypothetical scenario where k-means indicates 
that the optimal way to group the sequences would be into 3 groups: the four on the left would be group 1, the four 
sequences on the right would be group 3, and the rest are group 2. 
It is easy to add a fourth column to each of the dataframes of the class dendro object with that
information and then overlay the grouping on the dendrogram.

First, we use `mutate()` to assign a group to each sequence label:

```{r ggd3 mutate labels}

# Use mutate to add a column named Group which contains which group each sequence belongs to
# We can use the positions of the labels to parse that; labels with position along the x axis smaller than 5 belong to
# the 4 sequences on the left so we mark them as group 1, will do the same for the other two groups. 
# Can also use case_when.

hcggd$labels <- hcggd$labels %>%
  mutate(Group = if_else(x < 5, "1", "2")) %>%
  mutate(Group = if_else(x > 11, "3", Group))

```

Next, we apply a similar process to the segments dataframe:

```{r ggd3 mutate segments}

# Use a similar process with the segments
hcggd$segments <- hcggd$segments %>%  
  mutate(Group = if_else(x < 5, "1", "2")) %>%
  mutate(Group = if_else(x > 11, "3", Group))

```

Finally, we plot the grouped dendrogram:

```{r ggd3 plot}

# Now let's plot the dendrogram
p2 <- ggplot(data = hcggd$segments) +
  # Note that I color by the Group column
  # alpha denotes how transparent the coloring will be; the default is 1 and I want the colors to be
  # more gentle
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend, color=Group), 
               linewidth = 1.25, alpha=0.5) + 
  geom_text(data = hcggd$labels, aes(x, y-0.01, label = label, color = Group),
            show.legend = F,
            hjust = 1, angle = 90, size = 4, alpha=0.5)+
  theme_bw(base_size = 16)+
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = c(0.93,0.89),
        legend.background = element_blank())+
  # Setting the colors manually because the red and blue defaults are ugly
  scale_color_manual(values = c("dark red", "black", "dark green"))+
  ylim(min(hcggd$segments$y)-1, max(hcggd$segments$yend))+
  xlab(NULL)+
  ylab(NULL)
p2


```

<br>

***

::: {.callout-note}
**A note about machine learning**
While this plot might not look impressive, what you have done here is to take your first steps
(unless you've done this before) in machine learning. When you ask an algorithm (e.g., k-means
clustering) to classify your samples into groups, without specifying groups, you are performing
unsupervised machine learning. 
Once you have enough information about how the particular set of data classifies, you can use it to
develop a predictive algorithm that would classify new samples into groups based on the information
resulting from the unsupervised learning. When you train or use an algorithm to predict which
category a sample would be classified into, you are performing supervised machine learning.
:::


***

<br>


## **Trees**

Phylogenetic trees are tree-like graphs that not only denote relationships between each leaf on the
x axis, but also include a time component on the y axis (i.e., junctions lower on the tree are
assumed to have occurred earlier than those showing higher on the tree). Phylogenetic trees require 
storing more information than dendrograms (e.g., information about taxa, molecular information, 
genetic codes, model assumptions, etc.). The package ape offers the class phylo, which stores 
phylogenetic trees.

There are many methods to generate phylogenetic trees and several file formats in which
phylogenetic trees are written ([nexus](https://en.wikipedia.org/wiki/Nexus_file),
[newick](https://en.wikipedia.org/wiki/Newick_format) are the most common file formats). We will
go over those during lab 7. However, at the end of the day they all result in a phylo or phylo-like class of
variable. Here we will only get familiarized with said class. 

Because all tree-like objects in R are stored as lists, it is easy to convert between
them. The snippet below uses the `as.phylo()` function to generate a phylo object. 

```{r phylo1}

hcp <- ape::as.phylo(hc)
hcp

```

***

**Quick R exercise**

1. Explore the list outputted by `as.phylo()` (hcp). Which elements make up this list?

2. What class does each element belong to?

***

As you found out, `hcp` contains 4 elements. The matrix edge consists of two columns:

```{r phylo2}
#| echo: false

head(hcp$edge)

```

Each row in the matrix represents a branch or an edge. If this sounds like terminology from graph
theory, you are correct. Indeed, ape treats trees almost like association networks and actually
allows you to add columns to the edge matrix and plot a network.

The numeric vector edge.length contains the length of each branch of the tree:

```{r phylo3}
#| echo: false
head(hcp$edge.length)

```

Its order corresponds to the order of rows in the edge matrix.

The character vector `tip.label` surprisingly contains the label of each leaf of the tree, making
those easy to modify (for example, remove the UniRef90_ tag).

Nnode represents the number of nodes in the tree; again, this is terminology borrowed from graph 
theory and represents each branching point along the tree (in our case, 14).

As I mentioned, phylo class objects can contain additional information about the tree. The phylo object can include additional elements such as: 

* node.label: a character vector that contains node labels, for example bootstrap values.

* root.edge: a numeric value giving the length of the branch at the root if it exists.

* tip.data: a vector dating the tips (leaves) of the tree, useful if some are not current.




### **Plotting trees**

Plotting phylogenetic trees is straightforward using `baseR`. 

```{r basep}

plot(hcp)


```


Alternatively, trees can be unrooted.

```{r basep2}
plot(hcp, type = "unrooted")

```



Plotting phylogenetic trees using ggplot2 requires converting the phylo object to dendro and 
following the workflow we used earlier.

***

# **Summary assignment**

Start a script file in RStudio; name it 3315_lab4_yourinitials.R
You will submit this script and output file if applicable to Brightspace.

**1.** Convert the distance matrices generated for the dune dataset (euclidean and both Bray-Curtis) to an hclust object.

**2.** Convert the distance matrices generated for the dune dataset to an R matrix (you should have 3 matrices) and save them as CSV files. Make sure the file names are descriptive.

**3.** Convert the class hclust variables you have created to class dendrogram.

**4.** Plot both the class dendrogram and class hclust variables you created for the dune dataset using base R; give titles using the argument main. Save plots as jpeg using the RStudio plots tab.

  + Do you see any difference in clustering between each of the distance measures?

**5.** Convert the dendrogram objects to dendro objects and plot using `ggplot2`.

**6.** Export plots as pdf using `ggsave()`.

**7.** Add the column altitude to the label dataframe that's part of the dendro class variables; have the values "low", "high", "sea level" added to each sample randomly.

**8.** Add the column Groups to the segment dataframe of the dendro class variables you generated. 
Divide the two Bray-Curtis based dendrograms into 3 groups and the euclidean into 2 groups based on the
order of the samples.

**9.** Plot each of the dendro class objects; color the labels based on altitude and the branches based on the groups column. If this was real data, would you see any trends in clustering of these samples based on altitude? Make sure to export these plots as pdfs as well.

**10.** Use `?write.tree()` and `?write.nexus()` to export the phylogenetic tree generated in this lab. Give a descriptive file name so that it is clear which is a nexus and which is a newick format.

  + Which functions would you have used to import each of the files you exported?
  
**11.** Write a function that reads all csv files from a folder, log transforms them, and uses `metaMDS()` to perform NMDS on all files with binary bray as distance. Make sure that the results are all stored as elements of a list (hint: make sure that the list has the same number of elements as the number of files in the folder). Make sure that your function returns said list.

***