
# **Lab 5 population genetics and selection in R**{-}


**Learning objectives** 

__By the end of this lab you will be able to:__


*	Collaborate writing code

* Catch user input error

* Control flow using break and next

* Write custom functions performing statistical analysis

* Write functions modeling selection scenarios 

* Analyze population data

</details>


## **The concept**{-}

As bioinformaticians you will seldom work alone in your own brilliance. Most likely you will be a 
part of a team developing tools/analysis pipelines for other scientists, or have to explain your 
work to peers who are not bioinformaticians. Since we will not be using any new packages, this
seems like a good change to develop team working skill to to gain some experience in tool 
development.
During this lab you will work in groups and will write functions designed to analyse population 
data.

Before you begin solving problems, since these functions will be written with the end user in mind,
we will go over ways to catch user input errors and ways to control the flow of a function when a 
solution is reached.

### **Catching user input error**{-}

Sometimes, despite your best intentions, users will enter values to arguments which will result in
your code crushing or will send R in to a tail spin (e.g. vector with negative values when log
transforming). 

R has a lot of tools for "defensive programming", I will go over two: `stop()` and `message()`.
`stop()`, as its name implies, terminates the execution of the code if a deadly error was identified. In addition it prints an error message.

`message()`, allows you to print a customized message alerting the user to whatever it is you choose to altrt them about.

Let's look at how to use `stop()` and `message()` to catch user error.

```{r catch user error1}

#Let's write a very simple and useless function that takes a numeric vector 
#sums it up and calculate the log of the sum

Log_of_sum <- function(x){
#If the user inputs anything that is not a numeric vector the function will 
#not be able to run properly
#It is possible to test whether the input is in the correct format using an if 
#statement and an is.x function 
  
#If augment x is not numeric stop and print an error message explaining why was
#the execution was stopped
  
  if(!is.numeric(x)){
    stop("x is not a numeric vector")
  }
  
#Remember that matrices are vectors too, therefore is.numeric will result 
#in true for a numeric matrix as well
  
  if(is.matrix(x)){
    
#here we just send the user a message telling them the matrix was coerced to 
#a one dimensional vector
    
    message("x was coerced to a one dimensional vector")
    x <- c(x)
    
  }
  
#as always make sure your function returns a variable
  
  return(los <- log1p((sum(x))))
}


```

Now - _drum roll_ - let's use the function and see the magic that is stop. 

```{r catch user error2, error=TRUE}

#First generate a character vector

cv <- letters[1:5]

Log_of_sum(cv)

```

Note the error message. 

Now lets generate a matrix and get a message that it was coerced to a vector

```{r catch user error3}
#Why am I setting seed?
set.seed(1)

#Matrix with 2 columns 
m <- matrix(sample(1:5000,10), ncol = 2)

Log_of_sum(m)

```

#### **Quick R exercise - catching errors**{-}

A list is also a vector, what will happen if a list is used as a value for argument x in the 
`Log_of_sum()` function?

### **Control of flow the break and next functions with-in for loops**{-}

Sometimes, it is required to run a loop until certain condition (other than the loop
condition) is met. However, it is not always known when will this condition be met. For example,
when analyzing allele tables you have no way of knowing _a-priori_ at which generation will the
population no longer be in equilibrium for a certain allele. The function `break()` will terminate the
operation of a loop once an internal condition is satisfied.

Alternatively, it will be required to skip an iteration step if a certain condition was not met. If
we use the previous example, there is no way of knowing _a-priori_ if a population is no longer in HW
equilibrium. The function `next()` will skip iteration steps until a condition is met. 

The following code snippet demonstrates how to use `break()` and `next()` in a loop, within the context of
a function environment. Note that `break()` will terminate a loop and will exit to a higher level
environment. Within a function it will exit to the function environment and not to the global
environment.

```{r break and next1}

#This function consists of two loops one prints something until it encounters an element equal to
#to or larger than 100 and the second skips printing something until it encounters an element larger
#than than or equal to 100. 

bn <- function(x) {
  
  #conditional statement to catch input errors
  if(!is.numeric(x)){
    stop("x must be numeric")
  }
  
  #Initializing the counting variable
  
  i=1
  #setting up the loop condition
  #Note that this loop will run first
  for (i in seq_along(x) ) {
    
    #Seting up the break condition 
    if( x[i] < 100) {
      print ("This is loop1")
    } else {
      #Note that break will terminate the loop when encountering the first element that violates
      #the condition. There could be other elements out there.
      print(paste("loop 1 was terminated after",i,"iterations", sep = " "))
      break
    }
    
  }
  
  #Initializing the counting variable for the second loop
  
  i=1
  #Setting up the loop condition
  for (i in seq_along(x)) {
    #Setting up the next condition
    if(x[i]< 100){
      
      #Note that next will skip any element that meets the criterion
      next
    }else{
      #If criterion not met, print something
    print("This is loop2")}
    
  }
  
}
```

Let's see what the output of this function be with this random vector (why am I using a randomly generated vector?).

```{r}
#Why am I setting seed?
set.seed(1)

#Call the function bn
bn(sample(1:110,60))
```

As mentioned, the `break()` function terminated the loop after the first element to violate the
condition was encountered (the 23rd element of the vector). How many elements are there that do not
violate the condition of loop 1?

```{r}
#Let's ask R how meany elements in the vector are smaller than 100

set.seed(1)
#Note at the use of square brakes, sub sample the vector for only elements smaller than 100

length(sample(1:110,60)[sample(1:110,60)<100])
```

As you can see, `break()` skipped **_~half_** of the elements that met the condition.


## **The assignment**{-}

Bellow are 3 problems you will have to solve in groups.
The idea is that you work together to find the solution to each problem.
Submit the scripts and any associated files as a group by Tue Nov 1^st^ 23:59.

Each problem is worth 5 points

* 1 point will go to commenting, make sure to comment properly each step in your function

* If there is an attempt to solve the problem but no solution was made you will receive 2 additional marks on top of the marks for properly commenting the function

* If a solution was reached you will receive four additional marks on top of the marks for properly commenting the function

* If you feel that any of the members did not contribute at all to the assignment do not include their name and they will receive 0.

* **Remember to include a code to load any external library you used in your functions (e.g. ggplot2)**

* **Remember to make sure to test for correct input by user and to use `next()` and `break()` where appropriate**


* For problems 2 and 3 assume population size is constant (i.e. does not change from generation to
generation)!


* Remember that Hardy Weinberg equation for 2 alleles is:

  + Where p is the frequency of A~1~ and q is the frequency of A~2~

$$p^2 + 2pq + q^2$$
  
* Remember that:

$$\begin{align*}p+q = 1 \\p = 1-q\end{align*}$$


### **Problem 1**{-}

Write a function that tests for Chi square goodness of fit.
The purpose of this function is to be called by the other functions you will write in this assignment, 
so keep testing for HW equilibrium in mind.

Remember that the formula for Chi square goodness of fit is:

$$\chi^2 = \sum_{i=1}^{k} \frac{(Obs - Exp)^2}{Exp}$$
<br>

Remember that the critical value needed to reject the null hypothesis that observed values do not
differ significantly from the expected values is dependent on the number of categories you sum.
This is expressed as degrees of freedom which is defined as:

$$ df = n-1 $$
Where n is the number of categories, in your case the number of genotypes.

The critical value necessary to reject the null hypothesis can be retrieved using R's qchisq 
function, you can call it within your function. 


```{r getting them pvalues}

#Critical value for Chi square goodness of fit test, the 0.95 percentile with 
#1 - 3 degrees of freedom
#Note at how the function receives the argument, this is the critical value for a p-value equal to or smaller than 
#0.05, i.e. 5% probability to receive a larger Chi square value by chance.

qchisq(0.95, df = 1:3)

```

Figure out how many arguments your function will receive,(e.g. expected value, observed values,
significance level etc..). Remember that the test requires number of individuals rather than
proportions/frequencies so plan to account for that with respect to user input. 
What type of variable will each argument accept, and how will you test for correct input by the user. Think about 
whether you want to assign a default value to any of the arguments.

The function should return a list, which at minimum stores the following elements: 

* The significance level 

* The Chi square value

* Number of degrees of freedom

* Whether the null hypothesis is rejected or not

Since you will have to use it in following functions think whether require an element that
can be easily accessed by a conditional statement to ask if the null hypothesis was rejected.


### **Problem 2**{-}

Write a function that simulates allele dynamics under positive selection of a recessive allele.

Remember that the formula to calculate the change in q under this type of selection is:

$$ \Delta{q} = \frac{pq^2s}{1+q^2s} $$

The purpose of this function is to run through a selection scenario given initial q frequency and
determine at what generation a population is no longer at HW equilibrium.

At minimum the function should receive the initial proportion of q, the selection coefficient of 
the recessive allele, significance level for Chi square test, number of generations, and the 
population size.

Think about What type of variable will each argument accept, and how will you test for correct
input by the user. Think about whether you want to assign a default value to any of the arguments
(e.g. set a default number of generations, or a default significance level for the Chi square test). 
Lastly, think about how will you integrate the call of the Chi square function you wrote within 
the function you are currently writing, given that it requires number of individuals rather than frequencies.


The function should return a list with the following elements at minimum:

* The initial frequency of q

* The selection coefficient of the recessive allele

* The population size

* A dataframe with the columns for number of generations, frequency of q, frequency of p

  + Feel free to add additional columns if you think it necessary

* The generation in which the population is no longer in HW equilibrium

* A line plot generated by ggplot2 showing the frequency of each allele over time

  + The plot should include a vertical line that indicates the generation at which the population is no longer in equilibrium
  
  + The plot should include all generation inputted by the user even those past equilibrium
  
  + The plot should contain a descriptive title as well as axis titles
  
* The function should also output (print) the number of generations it took for the HW equilibrium to be
disrupted (with a descriptive message)

* The function should also output the plot generated by ggplot

**Run this function with s = 0.01 and the following population size: 100, 1000, 10000**

**Add this and the output plots to your submission** 

**Did your result change with population size? If so, how?**


### **Problem 3**{-}

Write a function that simulates allele dynamics under positive selection of co-dominant alleles.
Remember that the formula to calculate the change in q under this type of selection is:

$$ \Delta{q} = \frac{pqs}{1+2qs} $$

The purpose of this function is to accept either a matrix or a dataframe with number of individuals
per genotype, and determine if the population is under HW equilibrium. If not, determine if the
allele dynamics fit the case of co-dominance. 

The columns of the input table will be generation, A~1~A~1~, A~1~A~2~, A~2~A~2~.
Make sure to leave a comment so that the user will know this. The function should also receive as arguments, 
the selection coefficient and the significance level for a Chi square test.

Think about What type of variable will each argument accept, and how will you test for correct
input by the user. Think about whether you want to assign a default value to any of the arguments
(e.g. default significance level). Lastly, think about how will you integrate the call of the Chi
square function you wrote within your function, as you will need to call it twice here (hint calculate population size from the input table).

The function should return a list with the following elements at minimum:

* The input table

* The selection coefficient

* The generation in which HW equilibrium was disrupted

  + If not, than state that it didn't
  
* The chi square value, the critical value, the degrees of freedom

* Whether this is a co-dominance selection scenario

The function should output (print) the following information:

* Whether or not the population is in HW equilibrium

* What generation was the equilibrium disrupted

* Does the allele dynamics follow a co-dominance pattern

**Download the population data from BrighSpace**

**Run through your function and report output**