```{r}
#| label: setup-lab6
#| include: false
knitr::opts_chunk$set(echo = TRUE, class.source = "fold-show")
options(knitr.duplicate.label = "allow")
```

# **Lab 6: Sequence Evolution** {-}

The purpose of this lab is for you to get familiarized with methodologies used in the analysis of sequence similarity. 

<br>

## **Learning objectives** {-}

<details>
  <summary>Click to expand</summary>

__By the end of this lab you will be able to:__

* Demonstrate the ability to load biological sequences into R
* Perform dot-plot alignment using AA and nucleotide sequences
* Analyze sequences for regions of homology based on dot-plot alignments
* Perform global and local alignments using dynamic programming algorithms
* Analyze regions of homology based on those alignments
* Retrieve alignment scores
* Export sequences to be used for homology search in BLAST
* Perform multiple sequence alignment using the `msa` package
* Print multiple sequence alignments and identify conserved areas and consensus sequences
* Analyze differences in alignments with a tanglegram

</details>

***

<br>

****

::: {.callout-note}
**Note 1: The assignment questions for this week are interspersed throughout the tutorial.**
**They are itemized in the table of contents, so make sure you answer all of them.**
:::


****

****

::: {.callout-note}
**Note 2: Install the new packages ASAP, as some might take considerable time and may require you to update many dependencies.**
**If prompted to install UTF8, do not install from binaries.**
:::


To complete this lab, you will need to install the following packages: 

1. `dendextend`, a package that allows you to compare dendrograms

2. `msa`, a package for multiple sequence alignments

3. `ggmsa`, to plot MSAs with ggplot2

```{r}
#| label: package_install
#| eval: false
#| echo: true

# Installing the package dendextend
install.packages("dendextend")

# Installing the package ggmsa
BiocManager::install("ggmsa")

# Installing the package msa from Bioconductor
BiocManager::install("msa")
```

***

<br>

```{r}
#| label: package_load
#| eval: false
#| echo: true

# lapply is part of the apply family of functions, one of those functions that act as a for loop
lapply(c("vegan","ape","ggdendro","seqinr",
         "Biostrings", "tidyverse","dendextend","msa","ggmsa"),library, character.only = TRUE)
```


```{r}
#| label: p2
#| echo: false
#| message: false
#| warning: false
#| results: hide

lapply(c("vegan","ape","ggdendro","seqinr",
         "Biostrings", "tidyverse","dendextend","msa","ggmsa"),library, character.only = TRUE)
```

## **Pairwise alignment**

Let's begin the lab by conducting pairwise (PW) sequence alignments. To do so, you will use two libraries you are already familiar 
with: `Biostrings` and `seqinr`. You will perform a few types of pairwise sequence alignments using several sequences.  

### **Dot-plot alignment**

A dot-plot alignment is a simple way to perform pairwise sequence alignment. 
The purpose of this exercise is to grant you firsthand experience working with
dot-plot alignments and to understand their advantages and disadvantages.

Let's examine how well dot-plot alignments work with nucleotide sequences. You will 
align three 16S rRNA genes. You will use a file that contains 200 Archaeal rRNA genes obtained from [RefSeq](https://www.ncbi.nlm.nih.gov/refseq/about/).

```{r}
#| label: dotplot_nuc1

# Read the fasta file with Archaeal rRNA sequences from GitHub
arch <- seqinr::read.fasta(file = "https://raw.githubusercontent.com/idohatam/Biol-3315-files/main/Arch.fa")
```

Since the dataset contains both 16S rRNA and 23S rRNA sequences, let's find elements in 
the list imported by `read.fasta()` that house 16S rRNA sequences. We will do that by 
asking which positions house vectors with fewer than 1700 elements (the size of 16S rRNA 
genes ranges between 1400–~1650 bp).

```{r}
#| label: dotplot_nuc2
# Note: using the function lapply to iterate over the list returned by read.fasta
which(unlist(base::lapply(arch, function(x) length(x) < 1700)))
```

Looks like there are quite a few 16S rRNA sequences in the dataset. Let's get the last 3 elements. Note that `arch` is a list; therefore, the use of `lapply()` makes it very convenient to grab all three sequences without recycling code or using a `for` loop. 
Note as well that `lapply()` returns a list and requires unlisting just for ease of use later on.  

```{r}
#| label: dotplot_nuc3

arch_16s <- unlist(lapply(arch[(length(arch)-2):length(arch)],
                          function(x) paste0(x,collapse = "")))
```

Now let's figure out how to work with the function `dotPlot()`.

```{r}
#| label: dotplot_nuc4
#| eval: false

?dotPlot()
```

<br>

***

### **Question 1**

A. What is the length of each sequence? Show your code (1 mark).

B. Run the `dotPlot()` function to compare the first and third sequences, and to compare 
the second and third sequences using the default settings. Could you identify an optimal 
alignment using the default settings? Show your code and attach the resulting figure 
from each alignment saved as PDF (you can export from the
plot pane) (3 marks). **Hint:** Note that `dotPlot()` only accepts character vectors where each element is a single character (e.g., `c("A", "T", "G", "C")`). You might want to explore using `str_split_1()` from the `stringr` package.

C. Run the same alignments using a smoothing window with a word size of 50 letters, stringency of 30, and make sure that the window progresses one position at a time. Could you identify an optimal
alignment in any of the cases? Is there any apparent feature in any of the alignments?
Show your code and attach the resulting figures as PDFs (3 marks).

***

<br>

Next, let's examine how well a dot-plot alignment will perform when inspecting amino acid (AA) sequences.
Remember that AA sequences use a 20-letter alphabet, which in theory should reduce spurious matching.

To do that, first load the selA.fa file (found on GitHub) with `read.fasta()` from `seqinr`. 

```{r}
#| label: dot_plotaa
# Load selA using the read.fasta function from seqinr
selA <- read.fasta(file = "https://raw.githubusercontent.com/idohatam/Biol-3315-files/main/selA.fa")

# What class is this variable?
class(selA)

# How will you use it in subsequent functions?
```

<br>

***

### **Question 2**

A. Use the `dotPlot()` function with default settings to compare the 1st and 5th sequences in selA. Can you identify an alignment? Show your code and include a PDF of the figure produced (2 marks).

B. Now, smooth the sequences by selecting a window size of 10, stringency of 5, and a slide of one position. Did it improve the alignment? Show your code and attach a PDF of the resulting figure (2 marks).

C. What happens if you change the window size to 60 and select a stringency of 12? What does it tell you about results obtained by this technique? Show your code and include a PDF of the figure produced (3 marks).

***
<br>

### **Global PW alignment with NWA**

It's time to switch it up and use some fancier methods and fancier variable classes—yes, it's time
for some `Biostrings` and dynamic programming algorithms. 

You will work with the [Needleman-Wunsch algorithm (NWA)](https://en.wikipedia.org/wiki/Needleman%E2%80%93Wunsch_algorithm) for global alignment. Remember: global alignments align sequences from end to end, attempting to align every position in both sequences.

First, it is necessary to convert the sequences used in the previous example to `Biostrings` class variables. 
Furthermore, remember that a sequence alignment using a dynamic programming algorithm receives an alignment score. 
Therefore, it is necessary to create a scoring matrix for the nucleotide sequences (what will you use to score AA 
alignments?).

```{r}
#| label: nwa_convert_nuc
# Convert the nucleotide sequences to StringSets
arch_strinS <- DNAStringSet(arch_16s)
```

Now, let's convert the AA sequences to an `AAStringSet`. The AA sequences are saved as a list; also, they are lowercase and the substitution matrices are uppercase, so we need to correct for it. We will chain a few `lapply` functions to iterate over the list. The inner `lapply` pastes the sequences as strings, the second `lapply` changes the case to upper, then we unlist the list into a vector with the same number of elements as the list. Each element is a string that represents a sequence.

```{r}
#| label: nwa_convert_aa
selAset <- AAStringSet(unlist(lapply((lapply(selA, base::paste, collapse = "")),
                                    stringr::str_to_upper), '[[',1))
```

Need to set a scoring matrix for the alignment of nucleotide sequences; note that the scoring matrix only accounts for matches vs. mismatches; gap penalties will be addressed as arguments in the alignment function. 

```{r}
#| label: nwa_scoring_mat
nucsubmat <- nucleotideSubstitutionMatrix(match = 1, mismatch = -1, type = "DNA")
```

```{r}
#| label: nwa_help
#| eval: false
# How would you load the protein substitution matrix?
?substitution.matrices

# Arguments for pairwiseAlignment
?pairwiseAlignment
```

Now it's your turn. Run the function `pairwiseAlignment()` on the same nucleotide sequences used for
dot-plot. Remember that you need to subset the `XStringSet`. Use -1 as the penalty for opening and
extending. Make sure to use the correct substitution matrix. Store the resulting 
`PairwiseAlignmentsSingleSubject` in a list (you should have two elements in the list).

Similarly, run `pairwiseAlignment()` for the AA sequences from the previous example. Use the following
scoring matrices: BLOSUM62, BLOSUM80, PAM120, PAM40. What do the numbering systems for these
matrices represent? Store these in a list (you should have 4 elements in the list).

Did you notice how much faster NWA is compared with a dot-plot alignment?

```{r}
#| label: NWA2
#| echo: false
#| results: hide
#| message: false
#| warning: false

ga <- pairwiseAlignment(arch_strinS[1],arch_strinS[3], 
                        substitutionMatrix=nucsubmat, 
                        gapOpening = -1, gapExtension = -1,
                        type = "global")
ga


ga2 <- pairwiseAlignment(selAset[1],selAset[2], 
                         substitutionMatrix="PAM40", gapOpening = 1,
                         gapExtension = 1,type = "global")

ga2
```

<br>

***
### **Question 3**

A. Show the code used for each alignment; make sure to comment (2 marks).

B. Print the nucleotide alignments. What was the score for each of them? Given that 16S rRNA genes are highly conserved, would you expect these alignments to have negative scores (2 marks)?

C. Print each of the AA alignments. Did the trend in scores agree between PAM and BLOSUM (1 mark)?


***

<br>

### **Local PW alignment with SWA**

Now let's do the same, but perform a local alignment using the [Smith-Waterman algorithm (SWA)](https://en.wikipedia.org/wiki/Smith%E2%80%93Waterman_algorithm). Remember: local alignments identify the best matching subsequence regions rather than aligning sequences end to end. As before, make sure to 
store your alignments in two lists.  

```{r}
#| label: SWA
#| echo: false
#| results: hide

la <- pairwiseAlignment(arch_strinS[1],arch_strinS[3], 
                        substitutionMatrix=nucsubmat, gapOpening = -1, gapExtension = -1,
                        type = "local")
la


la2 <- pairwiseAlignment(selAset[1],selAset[2], substitutionMatrix="PAM40", gapOpening = 1,
                         gapExtension = 1,type = "local")

la2
```
<br>

***

### **Question 4**

A. Show the code used for each alignment; make sure to comment (2 marks).

B. What was the score and length of the local alignment between Arch1 and Arch3? What positions in Arch1 and Arch3 were aligned (remember that XString is an S4 object and should be subsetted using @ and not $)? Show your code (3 marks).

C. Run the sequence for Arch1 in nucleotide BLAST. Is it actually an archaeal sequence? What species and genus did the top hit belong to? Include a screenshot of the description and graphic summary tabs. Tip: check the box for excluding uncultured organisms from the alignment (3 marks).

D. What were the lengths of the AA alignments? Did they change with different substitution matrices? Show your code (2 marks).

***

<br>

## **MSA**

In this part, you will perform multiple sequence alignment (MSA) with the HBA and HBB sequences. You will read these files directly from GitHub using the URLs provided in the code.
These files contain datasets of hemoglobin A and hemoglobin B from several organisms. 
As you can see, the datasets are quite messy—i.e., the headers are long and inconsistent. Furthermore, the HBB dataset contains two fewer organisms than HBA. You'll need to clean up these datasets in order to compare alignments with `dendextend`.

The snippet below shows how to clean up the datasets once loaded. 

**Important:** Ensure that the HBA datasets contain only the sequences that are also present in the HBB datasets (hint: think about subsetting based on names).

First, we read the files as `XStringSet` objects.

```{r}
#| label: msa1_read
# Read files as XStringSet
hba_n <- readDNAStringSet(filepath = "https://raw.githubusercontent.com/idohatam/Biol-3315-files/main/HBA.fasta")
hba_aa <- readAAStringSet(filepath = "https://raw.githubusercontent.com/idohatam/Biol-3315-files/main/HBA_aa.fasta")
```

Next, we clean up the names of the nucleotide sequences. Splitting the names based on a blank space results in a matrix; we choose the 2nd and 3rd columns, which include the genus and species names. We use `apply()` to iterate over the columns and get a vector of names in the format "Genus Species". We will use functions from the `stringr` package.

```{r}
#| label: msa1_clean_nuc
# Note that for cleanup I am using functions from stringr; make sure to load it
names(hba_n) <- apply(str_split_fixed(names(hba_n), 
                                      pattern = " ", 4)[,2:3], 1, base::paste,
                      collapse = " ")
```

Finally, we clean up the names of the amino acid sequences. Here, genus and species are surrounded by brackets `[]`. We can use that to retrieve them. Note that `\\` are used as escape characters since `[]` have meaning in R syntax.

```{r}
#| label: msa1_clean_aa
names(hba_aa) <- str_remove_all(str_split_fixed(names(hba_aa), pattern = "\\[", n=2)[,2], "\\]")
```

>The MUSCLE implementation is a bit finicky with respect to names, so make sure to run over the 
>naming scheme for all `StringSets`, also note that MUSCLE might be finicky with respect to number of seqs, if one of the files
have more than 40 keep only the sequences that overlap between both files.
Now that all the sets are clean, it's time to perform MSA.
The function `msa()` can use CLUSTALW, MUSCLE, CLUSTAL Omega, or CLUSTAL 2.0. You will only use MUSCLE 
for the sake of this lab to keep things short. You will use the default settings, which means 
BLOSUM62 for AA alignments.

Here are the functions that you will need to use for the next part:

```{r}
#| label: msa1b
#| eval: false
?msa()
?msaConvert()
?msaConsensusSequence()
?dist.alignment()
```

You can also use the msa package [reference manual](http://www.bioconductor.org/packages/release/bioc/manuals/msa/man/msa.pdf).

***

::: {.callout-note}
**Note: You will also need the HBB.fasta and HBB_aa.fasta files. You can source those as well from my `GitHub` page.**
:::


***

```{r}
#| label: msa2
#| echo: false
#| results: hide

#Load the nucleic acid and amino acid sequences of HB-A and HB-B
hba_n <- readDNAStringSet(filepath = "https://raw.githubusercontent.com/idohatam/Biol-3315-files/main/HBA.fasta")
hba_aa <- readAAStringSet(filepath = "https://raw.githubusercontent.com/idohatam/Biol-3315-files/main/HBA_aa.fasta")
hbb_n <- readDNAStringSet(filepath = "https://raw.githubusercontent.com/idohatam/Biol-3315-files/main/HBB.fasta")
hbb_aa <- readAAStringSet(filepath = "https://raw.githubusercontent.com/idohatam/Biol-3315-files/main/HBB_aa.fasta")


#Fix the names
names(hba_n) <- apply(str_split_fixed(names(hba_n), 
                                      pattern = " ",4)[,2:3],1,
                      base::paste, collapse = " ")


names(hba_aa) <- str_remove_all(str_split_fixed(names(hba_aa),pattern = "\\[", n=2)[,2],"\\]")

names(hbb_aa) <- str_remove_all(str_split_fixed(names(hbb_aa),pattern = "\\[", n=2)[,2],"\\]")

# Subsetting to keep only sequences present in all datasets
# First, subset HBA AA to keep only sequences that are in HBB AA
hba_aa <- hba_aa[which(names(hba_aa)%in%names(hbb_aa))]
# Then, subset HBA nucleotide to keep only sequences that are in HBB AA
hba_n <- hba_n[which(names(hba_n)%in%names(hbb_aa))]
# Finally, subset HBB AA to keep only sequences that are in HBA nucleotide
hbb_aa <- hbb_aa[which(names(hbb_aa)%in%names(hba_n))]

hba_aa <- hba_aa[which(names(hba_aa)%in%names(hba_n))]

# After your current name cleanup, add this to make names MUSCLE-friendly
# MUSCLE prefers short, simple names without spaces or special characters



# Alignment for clustalW
hba_n_msa_c <- msa(hba_n,method = "ClustalW")

hba_aa_msa_c <- msa(hba_aa,method = "ClustalW")
hbb_aa_msa_c <- msa(hbb_aa,method = "ClustalW")

#Alignment with muscle

hba_n_msa_m <- msa(hba_n,method = "Muscle")
hba_aa_msa_m <- msa(hba_aa, method = "Muscle")

hbb_aa_msa_m <- msa(hbb_aa,method = "Muscle")


hba_aa_sq_m <- msaConvert(hba_aa_msa_m,type="seqinr::alignment")
hba_aa_sq_c <- msaConvert(hba_aa_msa_c,type="seqinr::alignment")

hba_n_sq_m <- msaConvert(hba_n_msa_m,type="seqinr::alignment")
hba_n_sq_c <- msaConvert(hba_n_msa_c,type="seqinr::alignment")

hbb_aa_sq_m <- msaConvert(hbb_aa_msa_m,type="seqinr::alignment")

hba_aa_sq_m_dist <- seqinr::dist.alignment(hba_aa_sq_m, "identity")
hba_aa_sq_c_dist <- seqinr::dist.alignment(hba_aa_sq_m, "identity")

hba_n_sq_m_dist <- seqinr::dist.alignment(hba_n_sq_m, "identity")
hba_n_sq_c_dist <- seqinr::dist.alignment(hba_n_sq_m, "identity")

hbb_aa_sq_m_dist <- seqinr::dist.alignment(hbb_aa_sq_m, "identity")

hba_aa_sq_m_clust <- as.dendrogram(hclust(hba_aa_sq_m_dist, method = "average"))
hba_aa_sq_c_clust <- as.dendrogram(hclust(hba_aa_sq_c_dist, method = "average"))

hba_n_sq_m_clust <- as.dendrogram(hclust(hba_n_sq_m_dist, method = "average"))
hba_n_sq_c_clust <- as.dendrogram(hclust(hba_n_sq_c_dist, method = "average"))

hbb_aa_sq_m_clust <- as.dendrogram(hclust(hbb_aa_sq_m_dist, method = "average"))

print(hbb_aa_msa_m, show = "complete")
```

<br>

As you have learned in lecture, one advantage of performing multiple sequence alignment is the 
ability to identify conserved domains and consensus sequences.
You will use ggmsa to visualize only AA alignments; the nucleotide ones are too large to print.
Interfacing with ggmsa requires a bit of a workaround, which is demonstrated in the snippet below.

First, we convert the MSA to an `AAStringSet`. `ggmsa` accepts `Biostrings` class types (e.g., `XStringSet`, `AAMultipleAlignment`, etc.), and `XStringSet` variables are much more malleable than `MultipleAlignment` variables.

```{r}
#| label: ggmsa_convert
hba_aa_msa_ms <- AAStringSet(AAMultipleAlignment(hba_aa_msa_m))
```

Converting to a `Biostrings` element removes the consensus sequence; let's add it as the last sequence. To do so, we turn the consensus sequence into an `AAStringSet` and add it as an additional element to the `AAStringSet` generated above. We are using `gsub()` to substitute `?` which marks ambiguity with `*` since `?` is not part of the `AAStringSet` alphabet.

```{r}
#| label: ggmsa_consensus
hba_aa_msa_ms[length(hba_aa_msa_ms)+1] <- AAStringSet(
  gsub("\\?", "*", msaConsensusSequence(hba_aa_msa_m)))
```

Lastly, add the name for the new consensus sequence, check the output, and pull up the help file for `ggmsa()`.

```{r}
#| label: ggmsa_name
names(hba_aa_msa_ms)[length(hba_aa_msa_ms)] <- "Consensus"

# Looks good
hba_aa_msa_ms[length(hba_aa_msa_ms)]
```

```{r}
#| label: ggmsa_help
#| eval: false
# How to use the ggmsa function
?ggmsa
```

<br>

***

### **Question 5**

A. Generate an MSA for each of the `StringSets` for HBA and HBB. Show your code (1 mark).

B. Save each alignment to your computer; you will use them in the next lab. The best approach is to convert to a Biostrings alignment set, then to a StringSet. Attach these files to your submission (1 mark).

C. Print the consensus sequence for each alignment. Which one worked better: nucleotide or AA? Show your 
code (2 marks).

D. Use ggmsa to print the AA alignments; print 50 positions at a time. Use `Chemistry_AA` as the color argument. For each protein, identify one position where there is a radical amino acid change (e.g., a change that significantly alters chemical properties such as charge, polarity, or size). Explain why this change is considered radical. Show your code and attach the figures as PDFs (use ggsave) (3 marks).

E. Convert the alignments to seqinr format and generate distance matrices. Show your code (2 
marks).

F. Cluster the matrices using UPGMA and convert to dendrograms. Show your code (2 marks).


***

<br>

### **Dendextend**

Lastly, let's compare the dendrograms using a tanglegram. Tanglegrams visually compare two dendrograms by connecting corresponding tips (leaves with matching labels), allowing you to assess whether the topologies differ. You can now compare if the alignments cluster differently based on whether nucleotides or amino acids are used. You can also inspect the topology between HBA and HBB.

```{r}
#| label: dext
#| out.width: "205%"

# Tanglegrams can be very colorful; the settings will remove color. Feel free to change.
# This one compares the alignments of HBA with CLUSTALW and MUSCLE; as you can see, they perform equally
# well on small datasets.
tanglegram(hba_aa_sq_c_clust,hba_aa_sq_m_clust, sort = TRUE, 
           common_subtrees_color_lines = FALSE, 
           highlight_distinct_edges  = FALSE, highlight_branches_lwd = FALSE)
```

***

<br>

### **Question 6**

A. Generate a tanglegram comparing the HBA alignment done with nucleotides and with amino acids. Did you find any
discrepancies with respect to membership in monophyletic groups? Show your code and attach the
figures (2 marks).

B. Perform the same analysis as in 6A for HBB (2 marks).

C. Generate a tanglegram comparing the alignments done on AA sequences for HBA and HBB. Did you find any discrepancies with respect to membership in monophyletic groups? Show your code and attach the
figures (2 marks). Keep this in mind when we study actual gene trees.

***
