
```{r setup lab7, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, class.source = "fold-show")
options(knitr.duplicate.label = "allow")
```


# **Lab 7 Reconstructing phylogenetic trees (gene trees) in R** {-}

**Learning objectives** 
__By the end of this lab you will be able to:__


*	Demonstrate the ability to load msa files into R

* Analyze and reconstruct phylogeny based on genes trees

* Identify and root tree using an out-groups

* Root trees using midpoint rooting

* Interface with NCBI and download taxonomic information

* Asses the confidence of your reconstruction using bootstrap

* Optimize and select substitution models for maximum likelihood phylogeny reconstruction

* Compare the reconstruction of phylogeny based different genes

* Compare the accuracy of different methodologies for tree reconstruction and inference

## **Using sequences to reconstruct phylogenetic trees - the setup** {-}


During this week's lab you will continue studying the the evolutionary relationships 
between HBA  and HBB sequences. You will generate multiple phylogenetic trees (gene trees)
and asses their quality.

****

>**Note1: The assignment questions for this week are interspersed throughout the tutorial.**
>**They are itemized in the table of content so make sure you answer all of them.**
>**The due date for the assignment will be announced by the lab instructor**

****

In order to complete this week's lab you will need to install the `phangorn` package winch extends
some of the capabilities of the `ape` package, especially with respect to incorporating substitution 
models. You will also need to install the package `phytools` which as it names suggests has some 
useful tools to manipulate phylogenetic trees. Lastly, you will need to download the package `taxize`
which allows to download the common names of the organisms from NCBI.

****

>**Note2: Install the new packages ASAP as some might take some time and would require you to update** 
>**a lot of dependencies**

****


```{r packages_lab7, eval = FALSE, echo = TRUE}

#Installing the package dendextend
install.packages("phangorn")

#Installing the package phytools
install.packages("phytools")

#Installing the package taxize
install.packages("taxize")


#lapply is part of the apply family of functions, one of those functions that act as a for loop
lapply(c("ape","seqinr","Biostrings", "tidyverse","msa", "phangorn",
         "phytools","stringr","taxize"),
       library, character.only = TRUE)

```

```{r p2, echo=FALSE,message=FALSE,warning=FALSE, results='hide'}

lapply(c("ape","seqinr","Biostrings", "tidyverse","msa", "phangorn",
         "phytools","stringr","taxize"),
       library, character.only = TRUE)
```

### **Import last week's MSA**{-}

Last week you have generated multiple sequence alignment for the HBA and HBB gene. This week, you 
will use the **amino acid** to generate phylogenetic trees. In order to do that first you will need to import them. 
You can import the msa using the read.phyDat function from pronghorn.
phyDat is a list based element used to stor_e alignment information by phangorn.  

```{r align_import, eval=FALSE}

#This is an example of how to read files using read.phyDat
hba_phy <- read.phyDat("hba_align_aa.fa", type = "AA")

```

<br>

### **Alternative, generate MSA**{-}

If you do not have the alingments from last week saved as a fasta file, or you still have 
the `msa` object on your worktable, you can use msaConvert function to convert the `msa` 
to `phyDat`. If you have the `msa` objects in your environment you can skip to the last 
step

```{r align_import2}
#Read file and align

hba_aa <- readAAStringSet(
  filepath = "https://raw.githubusercontent.com/idohatam/Biol-3315-files/main/HBA_aa.fasta")

hbb_aa <- readAAStringSet(
  filepath = "https://raw.githubusercontent.com/idohatam/Biol-3315-files/main/HBB_aa.fasta")

#Remember need to clean up names

names(hba_aa) <- str_remove_all(str_split_fixed(names(hba_aa),pattern = "\\[", n=2)[,2],"\\]")

names(hbb_aa) <- str_remove_all(str_split_fixed(names(hbb_aa),pattern = "\\[", n=2)[,2],"\\]")

#And make sure both have the same organisms
hba_aa <- hba_aa[which(names(hbb_aa)%in%names(hba_aa))]

#generate msa

hba_aa_msa <- msa(hba_aa, method = "ClustalW")

#Convert to a seqinr alignment and then th a phyDat class element

hba_phy <- as.phyDat(msaConvert(hba_aa_msa, "seqinr::alignment"),type = "AA")

```

<br>

## **Constructing a neighbor joining tree (distance based reconstruction)**{-}

One approach to reconstructing phylogenetic tree is by clustering OTUs based on distance (i.e. 
forming a distance matrix). 
There are two main algorithms to constructing a distance based tree, UPGMA and 
Neighbor joining (NJ). UPGMA assumes similar and constant rate of evolution across all OTUs, therefore 
the inference of evolutionary relations is error prone. NJ is more reliable will still being 
computationally light, since it does not make assumptions about evolutionary rates. To generate the 
distance matrix used to construct the tree you will use the dist.ml function from `phangorn`, which 
accounts for substitution mode. Since you are using an alignment of AA sequences you will need a 
model suitable for AA  substitution. You will use the Jones-Taylor-Thornton (JTT), an empirical model, to 
account for AA  substitution. Similar to BLOSUM but accounts for evolutionary history (i.e. is the observed number of 
substitutions the true number of substitution?).

```{r nj}

#First step is to generate a distance matrix using a substitution model.
#The function dist.ml from phangorn can use 17 different AA substitution models. 
#You will use a fairly simple one. 
hba_aa_dml <-dist.ml(hba_phy, model = "JTT")

#Next, let's generate a neighborjoining tree
#Note you will use NJ (capitalized) from phangoen rather than nj from ape to generate the tree
hba_aa_nj <- phangorn::NJ(hba_aa_dml)


#Note that the tree is unrooted
plot(hba_aa_nj)

```



Next let's root the tree. We do not have a good knowledge of the groups making up the tree yet, so you will use
midpoint rooting,which places the root midpoint between the two OTUs with the longest connecting path.
Use the `?` operator to figure out how to use the function `midpoint` and plot the rooted tree using the function `plot`.

```{r mp}
?midpoint

```

```{r mp2 ,echo=F, results='hide', fig.cap="The rooted tree"}
hba_aa_nj <- midpoint(hba_aa_nj)
plot(hba_aa_nj)
```

<br>

### **Bootstrapping the tree**{-}

There are multiple ways to root trees and there are multiple possible unrooted trees
that could be constructed when reconstructing phylogeny, however, just one of these trees is a true 
tree. While we have no way of knowing which is the true tree, we can
statistically asses the confidence in our tree topology. One way to asses this is by using
bootstrap which is an iterative algorithm to generate multiple random pseaudoreplicates of the tree
using the msa and assessing how many of them agree with the tree topology.
Bootstrapping requires a tree and the `phyDat` object generated at the beginning of the lab.

```{r bstrap}

#Use the bootstrap.phyDat function to calculate the bootstrap values of the tree
#Remember the bootstaping randominzes the columns of the alignment and generate pseaudoreplicates 
#of the tree.
#How would you make sure that the resultes of bootstapping are reproducible?

#here we are bootstraping the hba alignment it is important to choose the same model.
#A goos practice is to run bootstrap with at least 1000 iterations. You will run it 500 times, a 
#good compromise between accuracy and CPU power

hba_aa_nj_bs <-bootstrap.phyDat(hba_phy,
                                #note how the function is entered
                                FUN = function(x)NJ(dist.ml(x, model = "JTT")),bs=500)

#Note that values smaller than 50% are not printed.
plotBS(hba_aa_nj,hba_aa_nj_bs,"phylogram")


```

<br>

***

## **Question 1**{-}

A. Generate a bootstrapped neighbor joining tree for HBB and HBA. **Make sure that your code takes into account the randomization of the bootstrap method and allows reproducibility**. Show your code and make sure to comment (2 marks)

B. Print both trees, use the main argument and insert a descriptive plot title. Save the plots as PDFs and add them to your assignment. Show your code (2 marks).

C. The HBA tree has areas with bootstrap values lower than 50% do these areas have higher bootstrap values using HBB? What does it tell you about the resolving power of the HBs (2 marks)?


***

<br>

## **Constructing a maximum likelihood tree (character and distance base reconstruction)**{-}

Distance based approaches are rapid but considered quick and dirty since all they do is cluster 
based on distances. Maximum likelihood is an algorithm that calculates the likelihood that trees 
generated using a certain substitution models represent evolutionary events.

The function `modelTest` from the package `phangorn` allows to test which model is the best fit to construct the tree with a maximum likelihood algorithm. The output is a dataframe
with log likelihood and several estimators of prediction error (i.e. how accurate is the prediction that the model fits).
We will use Akaike information criterion (AIC). The test also offers a second estimator, Bayesian 
information criterion (BIC), which uses Bayesian inference. You can read more about them [here](https://machinelearningmastery.com/probabilistic-model-selection-measures/). 
In either case you will select the model with the lowest prediction error.

```{r ml}
#Let's use the modelTest function
mt <- modelTest(hba_phy, model=c("JTT", "LG", "WAG"))
#Print the results
mt
#Which model has the lowest AIC?
mt$Model[mt$AIC==min(mt$AIC)]


```

Looks like JTT is a good model to select, however, the test indicates the parameters G and I should be added to the 
model. G indicates gamma distribution, which assumes that the rate of evolution is unequal among OTUs and I indicates 
invariant sites i.e. assuming that some sites cannot be mutated (e.g. sites critically important to protein function).
Both parameters are reasonable considering that some areas are important to hemoglobin function and that mutation rate is different between different species.

Next let's use the model to construct a tree using maximum likelihood.

```{r ml2, results='hide'}
#Next let's use the NJ tree and the model selected by modelTest to rearange the tree topology such #that is represents the tree with the largest likelihood to be the true evolutionary tree.
#The output is a list like object of class pml.
#K is is the of intervals for gamma distribution
#inv is the proportion of inveriable sites

hba_aa_nj.pml <- pml(hba_aa_nj,hba_phy,model="JTT",k=4,inv=.2)

#Initially we set the parameters of the model, now that we have the tree let's optimize the parameters of the model with optim.pml this will take some time to complete.
hba_aa_nj.pml <- optim.pml(hba_aa_nj.pml,optNni=TRUE,optBf=TRUE,optQ=TRUE,optInv=TRUE,optGamma=TRUE,optEdge=TRUE)

```

**The tree generated by pml is unrooted, before you proceed make sure to root the tree with midpoint**

<br>

### **Bootstrap the maximum likelihood tree**{-}

Just like with NJ tree, we can asses the support of each of the edges.
The function `bootstrap.pml` performs bootstrap on ML trees. 

```{r mlb, results='hide', message=FALSE}
#Note that here we are only performing 100 itterations since more would take a lot of computational power

hba_aa_nj.pml.bs <- bootstrap.pml(hba_aa_nj.pml,bs=100,trees=TRUE,optNni=TRUE)

#Can plot the tree
plotBS(hba_aa_nj.pml$tree, hba_aa_nj.pml.bs, type = "phylogram")

```

***

<br>

## **Question 2**{-}

A. Generate a bootstrapped neighbor joining tree for HBB and HBA. **Make sure that your code takes into account the randomization of the bootstrap method and allows reproducibility**. Show your code and make sure to comment (2 marks)

B. Print both trees, use the main argument and insert a descriptive plot title. Save the plots as PDFs and add them to your assignment. Show your code (2 marks).

C. Compare the bootstrap values of the ML trees to the NJ trees, do ML bootstrap values offer better support to branches of the tree?


***

<br>

## **Comparing trees**{-}

Last, lets compare the trees that you generated today. You have already used the function 
`tanglegram` from `Dendextend` to compare the topology of dendrograms. The package 
`phytools` offers the function `cophylo` which allows for a similar 
comparison of phylogenetic trees.

But first...curious about what animals we have been working with? let's find out.

```{r taxize, results='hide', message=F}
#We are going to modefy the tips of the trees, so lets copy them just incase you need them again

hba_aa_nj2 <- hba_aa_nj

hba_aa_nj.pml2 <- hba_aa_nj.pml

#Next lets get a species names, they are the tip lables
#The order is important so we will get it for each tree

sp1 <- hba_aa_nj2$tip.label
sp2 <- hba_aa_nj.pml2$tree$tip.label

#Now lets get the species ids from ncbi

uids <- taxize::get_uid(sp1)

uids2 <- taxize::get_uid(sp2)

#Next let's get the common names of the organisms

cnames <- sci2comm(uids, "ncbi")
cnames2 <- sci2comm(uids2, "ncbi")

#The reterned element is a list, let's make is into a vector

a <- unlist(lapply(cnames, "[",1), use.names = F)

a2 <- unlist(lapply(cnames2, "[",1), use.names = F)

```

Before we continue, let's see if we have all the common names or if we have some missing 
values. Missing values will normally be denoted as NAs and they can mess with downstream 
processes. As bioinformaticians you will have to inspect data all the time for missing 
values and decide how to handle it. Here we will just use the scientific names. 
R has several functions to check for missing values. For 1d objects is.na is great, but 
for elements with 2d you can use the function complete.cases from the stats package which 
is part of BaseR.

```{r comnames}
#Are there any NAs?
which(is.na(a))
#Looks like we have 3 missing values, we will complete them from the original species list
#note how I subset the relevant values
a[which(is.na(a))] <- sp1[which(is.na(a))]

#Now for the ML tree as well
#Remember, the order of the tips on the tree might be different and that's why we can't use only #one list
a2[which(is.na(a2))] <- sp2[which(is.na(a2))]

a2

#Now let's change the tips of the trees

hba_aa_nj2$tip.label <- a
hba_aa_nj.pml2$tree$tip.label <- a2



```

***

Finally lets compare trees with `cophylo`.


```{r cophy}

#Here I am comparing the NJ and ML trees of HBA
#Note how I access the ML tree
obj <- cophylo(hba_aa_nj2, hba_aa_nj.pml2$tree)

#Plotting the trees
par(mfrow=c(1,1))
#mar sets the margins of the plot
plot(obj,mar=c(.1,.1,2,.1))
#Title is self explenatory
title("NJ                                                                   ML")
```

***

<br>

## **Question 3**{-}

A. Compare the following trees using `cophylo`: NJ and ML of HBA, NJ and ML of HBB, ML of 
HBA and ML of HBB (three comparisons in total). Show your code save the trees as PDF files
and attach to your assignment (2 marks).

B. Are there any clades that changed between HBA and HBB using ML (2 marks)?

C. Are there any clades that changed between NJ and ML, does it significantly change the
tree (i.e. do you see mammals clustering with fish all of a sudden?) (2 marks)?

D. HBA trees denote some snakes as a possible out-group. What taxonomic group do they 
belong to (e.g. is it a mammal)? Are there any other members of that taxonomic group on 
that tree, what does it tell you about this group (e.g. is it really monophyletic)? (2 marks)