```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, class.source = "fold-show")
options(knitr.duplicate.label = "allow")
```

# **Lab 8 comparing phylogenetic trees made with genes and genomes**{-}

**Learning objectives** 
__By the end of this lab you will be able to:__


*	Reconstruct phylogeny based on genes and genomes

* Evaluate the performance of different substitution models and choose an optimal model for the data-set

* Analyze and compare phylogenies constructed by different algorithms and different models

* Analyze and compare phylogenies constructed by genes and genomes

* Test the robustness of the reconstruction method


## **Phylogeny reconstruction using genomes - the setup**{-}

This will be the culmination of many concepts covered in both lecture and lab.
You will build a very simple analysis pipeline, and analyze the  relationships between 
nine different strains of corona viruses. You will do so using both genes and whole genomes and
compare the end result.

You will use both pairwise and multiple sequence alignment to find homologous and build gene and genome trees.
You would compare the genome trees to the gene trees and see which of them produces a more evolutionary relevant tree.

<br> 

***

>**Note that for this lab we will use a linear multiple sequence aligner to align genomes.**
>**This is only possible because you will be working with small viral genomes, ~10kb.**
>**Aligning larger genomes requires a [graph aligner](https://en.wikipedia.org/wiki/Sequence_graph), we will discuss these in depth in Biol4315**

***

<br>


### **Setting up local blast data sets**{-}

First import the functions allowing you to run NCBI's blast+ via R.


```{r source func, eval=FALSE}

source("https://raw.githubusercontent.com/idohatam/blastinR/master/Blast_func.R")
source("https://raw.githubusercontent.com/idohatam/blastinR/master/make_blast_db.R")
source("https://raw.githubusercontent.com/idohatam/blastinR/master/retrieve_hit_seqs.R")

```

Next, after downloading the viral genomes, turn them into a blast databse

```{r blast db, eval=FALSE}
make_blast_db(infile = "PATH/TO/VIRAL/GENOMES.fasta", dbtype = "nucl" ,outfile = "OUTFILE")
```

Perform blast with tblastn using the spike protein seqs as query.
Results will be returned as a dataframe/tibble. 

```{r bnr, eval=FALSE}
blast <- blstinr(btype = "tblastn", dbase = "your database", qry = "spike_protein.fasta ")

```

Retrieve the first hit for each of the spike proteins.
To do so you will need to get the id of each query sequence (i.e. the header minus the `>`)

```{r retrive hits, eval=FALSE}

retrieve_hit_seqs(query_ids = "vector with query ids", 
                  blast_results = "dataframe with your blast results",
                  blastdb = "your database")

```

>**Note: if you wish to use your own computer for this exersize make sure to download and install the [blast+](https://ftp.ncbi.nlm.nih.gov/blast/executables/blast+/LATEST/) suite. You can then source the Check_blast_function, to check if the software suite was installed correctly. 

<br>


### **Tips for this lab**{-}

* The code for this lab should include all the workflow from loading the library all the way to 
generating a maximum likelihood tree.

* When performing the multiple sequence alignment use the default setting (i.e. CLUSTAL) to save time (implementation of MUSCLE in msa is a bit finicky).

  * Note that the multiple sequence alignment of the genomes can take up to 40 min

* Remember to convert msa objects to phyDat, you can do that with msaConvert for the nucleotide sequences but will have to be creative for the protein sequences.
  
* Use the phangorn package to generate the NJ tree, use JC69 and the model for genomes and JTT as the model for the spike proteins when generating the distance matrix.

* When bootstrapping the NJ tree use 1000 iterations.

* When testing the models for ML use model = "all", this will allow you to compare the simplest substitution model with the most sophisticated model. You will be able to determine whether JC69 is up for the challenge.
  
  * Note that the optimization of the parameters of the likelihood models can take time.

* When bootstrapping the ML tree use 100 iterations

* When plotting (or comparing) trees make sure they are rooted

<br>

***

>**Note: Aligning the genomes, as well as the construction and bootstraping of the ML trees are computationally intensive. It is recommended that you use the computers of the Mac lab**

***

<br>

```{r packages, eval = FALSE, echo = FALSE}

#Installing the package dendextend
install.packages("phangorn")

#Installing the package phytools
install.packages("phytools")

#Installing the package taxize
install.packages("taxize")


#lapply is part of the apply family of functions, one of those functions that act as a for loop
lapply(c("ape","seqinr","Biostrings", "tidyverse","msa", "phangorn",
         "phytools","stringr","taxize"),
       library, character.only = TRUE)

```

<br>

## **Question 1**

Write a script that downloads the fasta files containing the spike protein sequences and the genomes sequences of nine
corona-viruses and performs a phylogenetic analysis.

***

>Link to genomes: 
>
>https://raw.githubusercontent.com/idohatam/Biol-3315-files/main/SARS_MERS_coronavirus.raw_sequence.fasta
>
>Link to spike: 
>
>https://raw.githubusercontent.com/idohatam/Biol-3315-files/main/spike.fa
>
>**Note: your code should download/read the files directly from  github, you're not allowed to download them manually**

***

a. Make sure that your script includes loading all the necessary libraries

b. Make sure to go from MSA all the way to a tanglegram using the cophy function from phylotoolt.

c. Make sure to comment every step of your script

d. each step of the script will receive 1 mark plus 3 marks for commenting, make sure to include the lines for the plots required in questions 2.

Total marks 13

<br>

## **Question 2**

a. Attached a bootstrapped rooted NJ tree of the spike protein and the genomes. Make sure to give a title. Export the trees as PDFs (2 marks)

b. Attache the following tanglegrams: ML vs NJ for genomes, ML vs NJ for proteins, ML genomes vs ML proteins. Attache the plots as PDFs. (2 marks)

<br>


## **Question 3**

a. Based on the tanglegrams was there any difference between the gene trees and the genome trees? Did the clades 
within the tree make sense biologically (i.e. did they resolve based on the taxonomic groups they belong to) (3 
marks)?

b. Do you think that your tanglgram would look the same had you built the spike tree with nucleic acid seqs rather
than amino acids (3 marks)?

c. Which model was found to be optimal for the genomes based on maximum likelihood? In no more than three sentences,
what are the assumptions behind this model and how do they compare to JC-69 (4 marks)?

d. How did JC-69 perform compared to the model selected by the maximum likelihood algorithm? Based on what you learned
in class, why do you think this is (2 marks)?

```{r files, eval=FALSE,echo=FALSE}

spike <- readAAStringSet(
  "https://raw.githubusercontent.com/idohatam/Biol-3315-files/main/spike.fa")

genomes <- readDNAStringSet(
  "https://raw.githubusercontent.com/idohatam/Biol-3315-files/main/SARS_MERS_coronavirus.raw_sequence.fasta")
```

```{r align, echo=FALSE,eval=FALSE}
stt <- Sys.time()

alsp <- msa(spike)

alng <- msa(genomes)

edt <- Sys.time()

edt-stt

```
```{r tree, eval=FALSE, echo=F}
gpd <- msaConvert(alng, type = "phangorn::phyDat")

spd <- as.phyDat(msaConvert(alsp, "seqinr::alignment"),type = "AA")

gdist <- dist.ml(gpd, model = "JC69")
gtree <- phangorn::NJ(gdist)

sdist <- dist.ml(spd, model = "JTT")
stree <- phangorn::NJ(sdist)

gbs <- bootstrap.phyDat(gpd,FUN = function(x)NJ(dist.ml(x, model = "JC69")),bs=1000)

plot(midpoint(gtree))


sbs <- bootstrap.phyDat(spd,FUN = function(x)NJ(dist.ml(x, model = "JTT")),bs=1000)

plot(midpoint(gtree))

plot(midpoint(gtree))

plot(midpoint(stree))

```

```{r model, eval=FALSE, echo=FALSE}
stt <- Sys.time()

mt <- modelTest(gpd, model="all")
mt2 <- modelTest(spd, model="all")

edt <- Sys.time()

edt-stt

?modelTest
#Print the results
mt
#Which model has the lowest AIC?
mt$Model[mt$AIC==min(mt$AIC)]
mt2$Model[mt2$AIC==min(mt2$AIC)]
stt <- Sys.time()
g.pml <- pml(gtree,gpd,model="GTR",k=4)
gopt.pml <- optim.pml(g.pml,optNni=TRUE,optBf=TRUE,optQ=TRUE,optGamma=TRUE,optEdge=TRUE)
plot(midpoint(gopt.pml$tree))

s.pml <- pml(stree,spd,model="WAG",k=4)
sopt.pml <- optim.pml(s.pml,optNni=TRUE,optBf=TRUE,optQ=TRUE,optGamma=TRUE,optEdge=TRUE)

edt <- Sys.time()

edt-stt

bsg.pml <- bootstrap.pml(gopt.pml, bs=500,trees=TRUE,optNni=TRUE)

#Can plot the tree
plotBS(midpoint(gopt.pml$tree), bsg.pml, type = "phylogram")


```


```{r thegram, echo=FALSE, eval=FALSE}

obj <- cophylo(midpoint(gtree), midpoint(sopt.pml$tree))

#Plotting the trees
par(mfrow=c(1,1))
#mar sets the margins of the plot
plot(obj,mar=c(.1,.1,2,.1))


```

