# **Lab 8: Comparing Phylogenetic Trees Made with Genes and Genomes**

<br>

## **Learning objectives** 
<details>
  <summary>Click to expand</summary>
__By the end of this lab you will be able to:__


* Reconstruct phylogeny based on genes and genomes

* Evaluate the performance of different substitution models and choose an optimal model for the dataset

* Analyze and compare phylogenies constructed by different algorithms and different models

* Analyze and compare gene trees to genome trees

* Test the robustness of the reconstruction method

</details>

<br>

## **Phylogeny reconstruction using genomes**

This will be the culmination of many concepts covered in both lecture and lab.
You will build a very simple analysis pipeline and analyze the relationships between 
nine different strains of coronaviruses. You will do so using both genes and whole genomes and
compare the end results.

You will use both pairwise and multiple sequence alignment to find homologs and build gene and genome trees.
You will compare the genome trees to the gene trees and see which of them produces a more evolutionarily relevant tree.

<br> 

***

::: {.callout-note}
**Note that for this lab we will use a linear multiple sequence aligner to align genomes.**
**This is only possible because you will be working with small viral genomes, ~10 kb.**
**Aligning larger genomes requires a [graph aligner](https://en.wikipedia.org/wiki/Sequence_graph); we will discuss these in depth in Biol 4315.**
:::

<br>

## **Tips for completing the lab**

* The lab assignment is due on the date specified by your instructor.

* This lab is made up of 3 questions that will require you to use the workflow you used for labs 6 and 7.

* The code for this lab should include all the workflow from loading the libraries all the way to 
generating and plotting a maximum likelihood tree.

* When performing the multiple sequence alignment, use the default settings (i.e., `ClustalW`) to save on run time.

  * Note that the multiple sequence alignment of the genomes may take a long time; it took my M1 Mac ~7 min. This is by far the slowest step and may need to be run on the lab Mac computers. If you have an older computer, I recommend you complete the maximum likelihood part on the lab computers as well.

* Remember to convert `msa` objects to `phyDat`. You can do that with `msaConvert()` for the nucleotide sequences. For the protein sequences, use: `as.phyDat(msaConvert(your_msa_object, "seqinr::alignment"), type = "AA")`.
  
* Use the `phangorn` package to generate the NJ tree; use JC69 as the model for genomes and JTT as the model for the spike proteins when generating the distance matrix.

* When bootstrapping the NJ tree, use 1000 iterations.

* When testing the models for ML, use `model = "all"`. This will allow you to compare the simplest substitution model with the most sophisticated model. You will be able to determine whether JC69 is up for the challenge.
  
  * Note that the optimization of the parameters of the likelihood models can take time; on my M1 Mac, it was about 6 min.

* When plotting (or comparing) trees, make sure they are rooted. Use the `ggtree` package introduced in Lab 7 for all tree visualizations.

<br>

***

::: {.callout-note}
**Note: Aligning the genomes, as well as the construction and bootstrapping of the ML trees, are computationally intensive. It is recommended that you use the computers in the Mac lab. Plan accordingly and start early.**
:::


***

<br>

```{r}
#| label: packages
#| eval: false
#| echo: false

# lapply is part of the apply family of functions, one of those functions that act as a for loop
lapply(c("ape","seqinr","Biostrings", "tidyverse","msa", "phangorn",
         "phytools","stringr","taxize", "ggtree"),
       library, character.only = TRUE)

```

<br>

## **Question 1**

Write a script that downloads the FASTA files containing the spike protein sequences and the genome sequences of nine
coronaviruses and performs a complete phylogenetic analysis.

***

>Link to genomes: 
>
>https://raw.githubusercontent.com/idohatam/Biol-3315-files/main/SARS_MERS_coronavirus.raw_sequence.fasta
>
>Link to spike: 
>
>https://raw.githubusercontent.com/idohatam/Biol-3315-files/main/spike.fa
>
::: {.callout-note}
**Note: Your code should download/read the files directly from GitHub; you are not allowed to download them manually.**
:::

Your script should include the following steps (each step is worth 1 mark):

1. Loading all necessary libraries
2. Reading/downloading the genome and spike protein files from GitHub
3. Performing MSA on both datasets
4. Converting MSA objects to phyDat format
5. Constructing NJ trees with appropriate models
6. Bootstrapping the NJ trees (1000 iterations)
7. Model testing for ML using `model = "all"`
8. Constructing ML trees with optimal models
9. Bootstrapping the ML trees (100 iterations)
10. Creating tanglegrams using the `cophylo()` function from `phytools`

Additionally, 3 marks are awarded for comprehensive commenting throughout your script.

**Requirements:**

a. Make sure that your script includes loading all the necessary libraries.

b. Make sure to go from MSA all the way to a tanglegram using the `cophylo()` function from `phytools`.

c. Make sure to comment on every step of your script.

d. Each step of the script will receive 1 mark, plus a total of 3 marks for commenting. Make sure to include the lines for the plots required in Question 2.

**Total marks: 13**

<br>

## **Question 2**

a. Attach a bootstrapped rooted NJ tree of the spike protein and the genomes. Make sure to give each a descriptive title. Export the trees as PDFs (2 marks).

b. Attach the following tanglegrams: ML vs. NJ for genomes, ML vs. NJ for proteins, ML genomes vs. ML proteins. Attach the plots as PDFs (2 marks).

<br>

## **Question 3**

a. Based on the tanglegrams, was there any difference between the gene trees and the genome trees? Did the clades 
within the tree make sense biologically (i.e., did they resolve based on the taxonomic groups they belong to) (3 
marks)?

b. Do you think that your tanglegram would look the same had you built the spike tree with nucleic acid sequences rather
than amino acids (3 marks)?

c. Which model was found to be optimal for the genomes based on maximum likelihood? In no more than three sentences,
what are the assumptions behind this model, and how do they compare to JC69 (4 marks)?

d. How did JC69 perform compared to the model selected by the maximum likelihood algorithm (as measured by AIC and BIC values)? Based on what you learned
in class, why do you think this is (2 marks)?

```{r}
#| label: files
#| eval: false
#| echo: false

spike <- readAAStringSet(
  "https://raw.githubusercontent.com/idohatam/Biol-3315-files/main/spike.fa")

genomes <- readDNAStringSet(
  "https://raw.githubusercontent.com/idohatam/Biol-3315-files/main/SARS_MERS_coronavirus.raw_sequence.fasta")
```

```{r}
#| label: align
#| echo: false
#| eval: false

stt <- Sys.time()

alsp <- msa(spike)

alng <- msa(genomes)

edt <- Sys.time()

edt-stt

```
```{r}
#| label: tree
#| eval: false
#| echo: false

gpd <- msaConvert(alng, type = "phangorn::phyDat")

spd <- as.phyDat(msaConvert(alsp, "seqinr::alignment"),type = "AA")

gdist <- dist.ml(gpd, model = "JC69")
gtree <- phangorn::NJ(gdist)

sdist <- dist.ml(spd, model = "JTT")
stree <- phangorn::NJ(sdist)

gbs <- bootstrap.phyDat(gpd,FUN = function(x)NJ(dist.ml(x, model = "JC69")),bs=1000)

ggtree(midpoint(gtree)) + geom_tiplab() + theme_tree2() + ggtitle("Genome NJ Tree")


sbs <- bootstrap.phyDat(spd,FUN = function(x)NJ(dist.ml(x, model = "JTT")),bs=1000)

ggtree(midpoint(stree)) + geom_tiplab() + theme_tree2() + ggtitle("Spike NJ Tree")

```

```{r}
#| label: model
#| eval: false
#| echo: false

stt <- Sys.time()

mt <- modelTest(gpd, model="all")
mt2 <- modelTest(spd, model="all")

edt <- Sys.time()

edt-stt

?modelTest
# Print the results
mt
# Which model has the lowest AIC?
mt$Model[mt$AIC==min(mt$AIC)]
mt2$Model[mt2$AIC==min(mt2$AIC)]
stt <- Sys.time()
g.pml <- pml(gtree,gpd,model="GTR",k=4)
gopt.pml <- optim.pml(g.pml,optNni=TRUE,optBf=TRUE,optQ=TRUE,optGamma=TRUE,optEdge=TRUE)

ggtree(midpoint(gopt.pml$tree)) + geom_tiplab() + theme_tree2() + ggtitle("Genome ML Tree")

s.pml <- pml(stree,spd,model="WAG",k=4)
sopt.pml <- optim.pml(s.pml,optNni=TRUE,optBf=TRUE,optQ=TRUE,optGamma=TRUE,optEdge=TRUE)

edt <- Sys.time()

edt-stt

# Note: This step is computationally intensive
bsg.pml <- bootstrap.pml(gopt.pml, bs=500,trees=TRUE,optNni=TRUE)

# Can plot the tree
boot_tree_g <- plotBS(midpoint(gopt.pml$tree), bsg.pml, type = "none")
ggtree(boot_tree_g) + geom_tiplab() + geom_nodelab() + theme_tree2() + ggtitle("Genome ML Tree with Bootstrap")

```


```{r}
#| label: thegram
#| echo: false
#| eval: false

obj <- cophylo(midpoint(gtree), midpoint(sopt.pml$tree))

# Plotting the trees
par(mfrow=c(1,1))
# mar sets the margins of the plot
plot(obj,mar=c(.1,.1,2,.1))

```
